---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(devtools)
load_all()
library(fdrtool)
```

# The `mitey` package

<!-- badges: start -->
[![R-CMD-check](https://github.com/kylieainslie/mitey/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kylieainslie/mitey/actions/workflows/R-CMD-check.yaml)
<!--[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/your-package)](https://CRAN.R-project.org/package=your-package) -->
[![codecov](https://codecov.io/gh/kylieainslie/mitey/branch/main/graph/badge.svg)](https://codecov.io/gh/kylieainslie/mitey)
[![License](https://img.shields.io/badge/license-MIT-blue.svg)](./LICENSE)
![GitHub Stars](https://img.shields.io/github/stars/kylieainslie/mitey)
<!-- [![DOI](https://zenodo.org/badge/DOI/<DOI>.svg)](https://doi.org/<DOI>)  -->
<!--[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/mitey)](https://CRAN.R-project.org/package=mitey) -->
<!-- badges: end -->

The `mitey` package is a lightweight package designed originally as a companion to the analyses presented by [Ainsie et al. 2024](https://kylieainslie.github.io/mitey/articles/epidemiology_of_scabies.html) on scabies transmission. However, these methods are more widely applicable than in the context of scabies, thus the motivation behind creating the `mitey` package was twofold and also provides flexible, documented code for methods to estimate epidemiological quantities of interest. 

Currently, `mitey` includes methods to estimate a) the mean and standard deviation of the serial interval distribution using a maximum likelihood framework developed by [Vink et al. 2014](https://doi.org/10.1093/aje/kwu209) and b) the time-varying reproduction number using the method developed by [Walling and Lipsitch 2007](https://pmc.ncbi.nlm.nih.gov/articles/PMC1766383/).

### Estimating the serial interval
The method developed by Vink et al. uses data about the time of symptom onset with no precise information about transmission pairs and an assumed underlying serial interval distribution (either Gaussian or Gamma) to estimate the mean and standard deviation of the serial interval distribution. Briefly, the method involves calculating the index case-to-case (ICC) interval for
each person, where the person with the earliest date of symptom onset will be considered the index case. The rest of the individuals will have an ICC interval calculated as the number of days between their symptom onset and the index case. 

## Installation

1.  Install [R](http://cran.r-project.org)

2.  Install the development version of serosolver from
    [GitHub](https://github.com/kylieainslie/mitey):

``` r
# install.packages("devtools")
devtools::install_github("kylieainslie/mitey")
```

## Guide and vignettes

A quick start guide and detailed vignettes are still in development. 

Validation of the method used to estimate the mean and standard deviation of the serial interval proposed by [Vink et al. 2014](https://doi.org/10.1093/aje/kwu209) can be found [here](https://kylieainslie.github.io/mitey/articles/code_validation_for_Vink_method.html).

The manuscript pre-print that motivated the development of this package can be found [here](https://kylieainslie.github.io/mitey/articles/epidemiology_of_scabies.html).

Validation of the method used to estimate the time-varying reproduction number proposed by [Wallinga and Lipsitch 2007](https://pmc.ncbi.nlm.nih.gov/articles/PMC1766383/) can be found [here](https://kylieainslie.github.io/mitey/articles/rt_estimation_validation.html). *Note: this vignette is still under development, so is incomplete.*

## Example
This is a basic example of estimating the mean serial interval from data on time of symptom onset for an infectious disease. We first simulate index case-to-case (ICC) intervals. Here, we directly simulate the ICC intervals based on their route of transmission. The specified mean serial interval `hmu` is 15 and the specified standard deviation `hsigma` is 3. The weights for each route of transmission are specified as `hw1`, `hw2`, `hw3`, and `hw4`, respectively.

```{r load_packages, eval = FALSE}
library(mitey)
library(fdrtool)
```

```{r simulated_data}
set.seed(1234)

N <- 5000; hmu<-15; hsigma<-3; hw1 <- 0.2; hw2 <- 0.5; hw3 <- 0.2; hw4 <- 0.1

CP <- rhalfnorm((hw1*N),theta=sqrt(pi/2)/(sqrt(2)*hsigma))
PS <- rnorm(hw2*N,mean=hmu,sd=hsigma)
PT <- rnorm(hw3*N,mean=2*hmu,sd=sqrt(2)*hsigma)
PQ <- rnorm(hw4*N,mean=3*hmu,sd=sqrt(3)*hsigma)

sim_data <- round(c(CP,PS,PT,PQ))
```

We can look at our simulated data by plotting them as a histogram.
```{r plot_sim_data, echo=FALSE, fig.cap="Figure 1. Histogram of index-case to case intervals (in days) for simulated data."}
breaks<-seq(-.5,(max(sim_data)+1),by=1)
hist(sim_data,breaks=breaks,prob=TRUE,main="",xlab="Index-case to case interval (days)", ylab="frequency", ylim=c(0,0.07))
```

Then, we estimate the mean and standard deviation of the simulated serial interval using the `si_estim` function.
```{r si_estim_sim_data}
results <- si_estim(sim_data)
results
```
The output of `si_estim` is a named list with elements `mean`, `sd`, and `wts`, which contain the estimated mean, standard deviation, and weights of the serial interval distribution, respectively. We see that using the simulated data and assuming an underlying normal distribution, we obtain estimates very close to the input values: a mean serial interval estimate of `r round(results$mean[1], 2)` and a standard deviation of `r round(results$sd[1], 2)`. We are also able to recapture the input weights: hw1 = `r round(results$wts[1], 2)`, hw2 = `r round(results$wts[2] + results$wts[3], 2)`, hw3 = `r round(results$wts[4] + results$wts[5], 2)`, hw4 = `r round(results$wts[6] + results$wts[7], 2)`.

Using the `plot_si_fit` function, we can use the outputs of `si_estim` to plot the fitted serial interval over the symptom onset data.
```{r plot_sim_data_fit, fig.cap="Figure 2. Fitted serial interval curves plotted over symptom onset data for simulated symptom onset data. Red line is the fitted serial interval curves assuming an underlying Normal distribution."}
plot_si_fit(
    dat = sim_data,
    mean = results$mean[1],
    sd = results$sd[1],
    weights = c(results$wts[1], results$wts[2] + results$wts[3],
                results$wts[4] + results$wts[5], results$wts[6] + results$wts[7]),
    dist = "normal"
  )
```

<!--You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->
