---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# The `mitey` package

<!-- badges: start -->
<!-- badges: end -->

The motivation behind creating the `mitey` package was twofold: 1) to provide the data and code to reproduce the results in Ainslie, K, Hooiveld, M, and Wallinga, J. 2024. On the epidemiological characteristics of scabies. \textit{(in preparation)}, and 2) to provide flexible, documented code for methods not previously available in `R` that can help estimate epidemiological quantities of interest. 

## Installation

1.  Install [R](http://cran.r-project.org)

2.  Install the development version of serosolver from
    [GitHub](https://github.com/kylieainslie/mitey):

``` r
# install.packages("devtools")
devtools::install_github("kylieainslie/mitey")
```

## Guide and vignettes

A quick start guide and detailed vignettes are still in development. 

Validation of the method used to estimate the mean and standard deviation of the serial interval proposed by [Vink et al. 2014](https://doi.org/10.1093/aje/kwu209) can be found [here](https://kylieainslie.github.io/mitey/articles/code_validation_for_Vink_method.html).

The manuscript pre-print that motivated the development of this package can be found [here]().

## Example
This is a basic example of estimating the mean serial interval from data on time of symptom onset for an infectious disease. We first simulate index case-to-case (ICC) intervals. Here, we directly simulate the ICC intervals based on their route of transmission. The specified mean serial interval `hmu` is 15 and the specified standard deviation `hsigma` is 3. The weights for each route of transmission are specified as `hw1`, `hw2`, `hw3`, and `hw4`, respectively.

```{r load_packages}
library(mitey)
library(fdrtool)
```

```{r simulated_data}
set.seed(1234)

N <- 5000; hmu<-15; hsigma<-3; hw1 <- 0.2; hw2 <- 0.5; hw3 <- 0.2; hw4 <- 0.1

CP <- rhalfnorm((hw1*N),theta=sqrt(pi/2)/(sqrt(2)*hsigma))
PS <- rnorm(hw2*N,mean=hmu,sd=hsigma)
PT <- rnorm(hw3*N,mean=2*hmu,sd=sqrt(2)*hsigma)
PQ <- rnorm(hw4*N,mean=3*hmu,sd=sqrt(3)*hsigma)

sim_data <- round(c(CP,PS,PT,PQ))
```

We can look at our simulated data by plotting them as a histogram.
```{r plot_sim_data, echo=FALSE, fig.cap="Figure 1. Histogram of index-case to case intervals (in days) for simulated data."}
breaks<-seq(-.5,(max(sim_data)+1),by=1)
hist(sim_data,breaks=breaks,prob=TRUE,main="",xlab="Index-case to case interval (days)", ylab="frequency", ylim=c(0,0.07))
```

Then, we estimate the mean and standard deviation of the simulated serial interval using the `si_estim` function.
```{r si_estim_sim_data}
results <- si_estim(sim_data)
results
```
The output of `si_estim` is a named list with elements `mean`, `sd`, and `wts`, which contain the estimated mean, standard deviation, and weights of the serial interval distribution, respectively. We see that using the simulated data and assuming an underlying normal distribution, we obtain estimates very close to the input values: a mean serial interval estimate of `r round(results$mean[1], 2)` and a standard deviation of `r round(results$sd[1], 2)`. We are also able to recapture the input weights: hw1 = `r round(results$wts[1], 2)`, hw2 = `r round(results$wts[2] + results$wts[3], 2)`, hw3 = `r round(results$wts[4] + results$wts[5], 2)`, hw4 = `r round(results$wts[6] + results$wts[7], 2)`.

Using the `plot_si_fit` function, we can use the outputs of `si_estim` to plot the fitted serial interval over the symptom onset data.
```{r plot_sim_data_fit, fig.cap="Figure 2. Fitted serial interval curves plotted over symptom onset data for simulated symptom onset data. Red line is the fitted serial interval curves assuming an underlying Normal distribution."}
plot_si_fit(
    dat = sim_data,
    mean = results$mean[1],
    sd = results$sd[1],
    weights = c(results$wts[1], results$wts[2] + results$wts[3],
                results$wts[4] + results$wts[5], results$wts[6] + results$wts[7]),
    dist = "normal"
  )
```

<!--You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->
