---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# The `mitey` package

<!-- badges: start -->
[![R-CMD-check](https://github.com/kylieainslie/mitey/actions/workflows/R-CMD-Check.yaml/badge.svg)](https://github.com/kylieainslie/mitey/actions/workflows/R-CMD-Check.yaml)
[![License: EUPL v1.2](https://img.shields.io/badge/License-EUPL%20v1.2-blue.svg)](https://interoperable-europe.ec.europa.eu/collection/eupl/eupl-text-eupl-12)
![GitHub Stars](https://img.shields.io/github/stars/kylieainslie/mitey)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15446456.svg)](https://doi.org/10.5281/zenodo.15446456) 

<!-- [![CRAN status](https://www.r-pkg.org/badges/version/mitey)](https://CRAN.R-project.org/package=mitey) -->
<!-- [![CRAN downloads](https://cranlogs.r-pkg.org/badges/mitey)](https://cran.r-project.org/package=mitey) -->
<!-- badges: end -->

The `mitey` package is a lightweight package designed originally as a companion to the analyses presented by [Ainslie et al. 2025](http://dx.doi.org/10.2139/ssrn.5184990) on scabies transmission. However, the methods featured in `mitey` are more widely applicable than in the context of scabies. Thus, the motivation behind creating the `mitey` package was twofold: 1) provide reproducible code to reproduce  and also provides flexible, documented code for methods to estimate epidemiological quantities of interest. 

Currently, `mitey` includes methods to estimate a) the mean and standard deviation of the serial interval distribution using a maximum likelihood framework developed by [Vink et al. 2014](https://pubmed.ncbi.nlm.nih.gov/25294601/) and b) the time-varying reproduction number using the method developed by [Walling and Lipsitch 2007](https://pmc.ncbi.nlm.nih.gov/articles/PMC1766383/).

## Installation

Install `mitey` from [CRAN](https://cran.r-project.org):

```r
install.packages("mitey")
```

Or, install the development version of `mitey` from [GitHub](https://github.com/kylieainslie/mitey):

``` r
devtools::install_github("kylieainslie/mitey")
```
#### Installation time
```{r install_time}
system.time({
  devtools::install_github("kylieainslie/mitey", force = TRUE)
})
```

## Main Functions

| Function | Description |
|----------|-------------|
| `si_estim()` | Estimates the mean and standard deviation of the serial interval distribution |
| `plot_si_fit()` | Plots the fitted serial interval distribution |
| `wallinga_lipsitch()` | Estimates the time-varying reproduction number using the Wallinga & Lipsitch method |
| `generate_synthetic_epidemic()` | Generates a synthetic epidemic curve with specified parameters for testing and simulation |

## Example Usage

```{r example_si}
library(mitey)

#-----------------
# Serial Interval Estimation
#-----------------
icc_intervals <- c(rep(6,4),rep(7,8),rep(8,14),rep(9,31),rep(10,29),rep(11,42),rep(12,25),rep(13,16),rep(14,16), rep(15,10),rep(16,4),rep(17,2),rep(18,2))

# Estimate serial interval
si_results <- si_estim(icc_intervals)
si_results
```

```{r example_si_plot}
# Plot fitted serial interval distribution
plot_si_fit(
  dat = icc_intervals,
  mean = si_results$mean[1],
  sd = si_results$sd[1],
  weights = c(si_results$wts[1], 
              si_results$wts[2] + si_results$wts[3],
              si_results$wts[4] + si_results$wts[5], 
              si_results$wts[6] + si_results$wts[7]),
  dist = "normal"
)
```

```{r example_rt}
#-----------------
# Time-varying Reproduction Number Estimation
#-----------------
# Generate a synthetic epidemic using the generate_synthetic_epidemic function
set.seed(123)
true_r <- c(rep(2.5, 50), seq(2.5, 0.8, length.out = 30), rep(0.8, 100))

# Generate synthetic epidemic data
synthetic_data <- generate_synthetic_epidemic(
  true_r = true_r,      # Time-varying reproduction number
  si_mean = 5.2,        # Mean serial interval (days)
  si_sd = 1.72,         # SD of serial interval
  si_dist = "gamma",    # Distribution type
  initial_cases = 10    # Initial number of cases
)

incidence <- synthetic_data$incidence
dates <- synthetic_data$date
```

```{r example_rt_epicurve, echo=FALSE}
# Plot the epidemic curve
par(mar = c(4, 4, 2, 1))
plot(dates, incidence, type = "h", col = "darkred", lwd = 2,
     main = "Synthetic Epidemic Curve",
     xlab = "Date", ylab = "Daily Cases")
```

```{r example_rt_est}
# Estimate time-varying reproduction number 
results <- wallinga_lipsitch(
  incidence = incidence, 
  dates = dates,
  si_mean = 5.2,          # mean serial interval in days
  si_sd = 1.72,           # serial interval SD
  si_dist = "gamma",      # serial interval distribution
  smoothing = 7,          # 7-day smoothing window
  bootstrap = TRUE,       # calculate bootstrap confidence intervals
  n_bootstrap = 100,      # number of bootstrap samples
  conf_level = 0.95       # 95% confidence intervals
)

head(results)
```

```{r example_rt_plot, echo=FALSE}
# Plot the results with confidence intervals
par(mar = c(4, 4, 2, 1))
plot(dates, results$R, type = "l", lwd = 2,
     ylim = c(0, max(results$R_upper, na.rm = TRUE) + 0.2),
     xlab = "Date", ylab = "Reproduction Number (Rt)",
     main = "Time-varying Reproduction Number",
     col = "darkblue")

# Add confidence interval as shaded area
polygon(c(dates, rev(dates)), 
        c(results$R_lower, rev(results$R_upper)), 
        col = rgb(0, 0, 1, 0.2), border = NA)

# Add true R
lines(dates, true_r, lty=2, col = "black")
# Add legend
legend("topright", 
       legend = c("Estimated Rt", "95% CI", "True Rt"),
       col = c("darkblue", rgb(0, 0, 1, 0.2), "black"),
       lwd = c(2, 10, 2),
       lty = c(1, 1, 2),
       bty = "n")
```

## Expected Run Times
Expected run times for the examples provided above are shown in the table below.

<details>
<summary>Click to expand run time information</summary>

| Example | Data Size | Run Time | Details |
|---------|-----------|----------|---------|
| **Serial Interval Estimation** | | | |
| `si_estim()` | 203 ICC intervals | **< 1 second** | EM algorithm on example outbreak data |
| `plot_si_fit()` | Same data + plotting | **< 1 second** | Visualization of fitted distribution |
| **Synthetic Data Generation** | | | |
| `generate_synthetic_epidemic()` | 180 days | **< 1 second** | Creates epidemic curve with known Rt |
| **Reproduction Number Estimation** | | | |
| `wallinga_lipsitch()` (point estimates) | 180 days, no bootstrap | **1-2 seconds** | Fast point estimates only |
| `wallinga_lipsitch()` (with bootstrap) | 180 days, 100 bootstrap samples | **2-5 minutes** | Including 95% confidence intervals |
| **Complete Example Workflow** | | | |
| All example code blocks | Full workflow | **3-6 minutes** | Serial interval + Rt estimation + plots |

**Performance tip**: Set `bootstrap = FALSE` for quick demonstrations, or reduce `n_bootstrap` to 10-20 for faster approximate confidence intervals.
</details>

## Vignettes

- A quick start guide showing examples of how to estimate the serial interval and time-varying reproduction number can be found [here](https://kylieainslie.github.io/mitey/articles/quick_start_guide.html).

- A script that reproduces the results from [Ainslie et al. 2025](http://dx.doi.org/10.2139/ssrn.5184990) can be found [here](https://kylieainslie.github.io/mitey/articles/reproduce_results_ainslie_et_al.html).

- Validation of the method used to estimate the mean and standard deviation of the serial interval proposed by [Vink et al. 2014](https://pubmed.ncbi.nlm.nih.gov/25294601/) can be found [here](https://kylieainslie.github.io/mitey/articles/code_validation_for_Vink_method.html).

- Validation of the method used to estimate the time-varying reproduction number proposed by [Wallinga and Lipsitch 2007](https://pmc.ncbi.nlm.nih.gov/articles/PMC1766383/) can be found [here](https://kylieainslie.github.io/mitey/articles/rt_estimation_validation.html). 

## Data

Several data files are stored in the repo so that the results presented in [Ainslie et al. 2025](http://dx.doi.org/10.2139/ssrn.5184990) are reproducible. Data files are stored in `inst/extdata/data/`. Below is a brief description of the different files.

- `si_data.rds`
  - *Description:* Data on date of symptom onset for scabies outbreaks described by [Kaburi et al.](https://doi.org/10.1186/s12889-019-7085-6), [Akunzirwe et al.](https://doi.org/10.21203/rs.3.rs-3205380/v1), [Tjon-Kon-Fat et al.](https://doi.org/10.1371/journal.pntd.0009485), and [Ariza et al.](https://doi.org/10.1007/s10096-012-1752-1). For all outbreaks except Kaburi et al. the raw data was not available, thus the date of symptom onset data had to be reconstructed using the epidemic curve provided in the manuscript. The original data from Kaburi et al. is also available in the `data` directory (`Kaburi_et_al_data_scabies.xlsx`).

- `scabies_data_yearly.xlsx`
  - *Description:* Annual scabies incidence per 1000 people in the Netherlands from 2011-2023.
  - *Source:* [Nivel](https://www.nivel.nl/nl/zorg-en-ziekte-in-cijfers)
- `scabies_data_consultation_weekly.xslx`
  - *Description:* Weekly numbers of persons consulting for scabies (per 100,000 people) from 2011 to 2023 in the Netherlands as diagnosed by general practitioners (GPs). *Note:* Individuals in institutions (e.g., care homes, prisons) usually have their own health care provider and are generally not taken into account in GP registrations.
  - *Source:* [Nivel](https://www.nivel.nl/nl)

## License

This package is distributed under the European Union Public License (EUPL) v1.2. See LICENSE file for details.

## Citation

If you use this package, please cite both the manuscript and the software:

#### Manuscript
Ainslie, K.E.C., M. Hooiveld, J. Wallinga. (2025). Estimation of the epidemiological characteristics of scabies. *Available at SSRN*. https://papers.ssrn.com/sol3/papers.cfm?abstract_id=5184990

#### Software
```{r}
citation("mitey")
```

#### Other
For reference, here are the original methodological papers:

- Vink et al. (2014). Serial intervals of respiratory infectious diseases: A systematic review and analysis. American Journal of Epidemiology, 180(9), 865-875.
- Wallinga, J., & Lipsitch, M. (2007). How generation intervals shape the relationship between growth rates and reproductive numbers. Proceedings of the Royal Society B: Biological Sciences, 274(1609), 599-604.

## System Requirements

- *R Version:* R >= 4.0.0 (developed and tested with R 4.5.0)
- *Operating Systems:* Tested on 
  - macOS (latest via GitHub Actions)
  - Windows Server (latest via GitHub Actions)
  - Ubuntu Linux 22.04+ (with R release and R-devel)

#### Dependencies

<details>
<summary>Click to expand dependency information</summary>

| Category | Package | Purpose | Required |
|----------|---------|---------|----------|
| **Core Dependencies** | | | |
| | `fdrtool` | Half-normal distribution functions (Vink method) | ✅ Yes |
| | `stats` | Statistical distribution functions | ✅ Yes |
| | `brms` | Bayesian meta-analysis | ✅ Yes |
| **Data Manipulation** | | | |
| | `dplyr` | Data manipulation and grouping | 📦 Suggested |
| | `tidyr` | Data reshaping | 📦 Suggested |
| | `purrr` | Functional programming tools | 📦 Suggested |
| **Visualization** | | | |
| | `ggplot2` | Statistical graphics | 📦 Suggested |
| | `cowplot` | Combining plots | 📦 Suggested |
| | `ggridges` | Ridge plots for meta-analysis | 📦 Suggested |
| | `viridis` | Color palettes | 📦 Suggested |
| **Time Series & Data Handling** | | | |
| | `zoo` | Moving averages and smoothing | 📦 Suggested |
| | `lubridate` | Date manipulation | 📦 Suggested |
| | `ISOweek` | ISO week handling | 📦 Suggested |
| | `openxlsx` | Excel file reading | 📦 Suggested |
| **Method Validation** | | | |
| | `EpiEstim` | Reproduction number estimation comparison | 🔬 Development |
| | `EpiLPS` | Alternative reproduction number methods | 🔬 Development |
| | `outbreaks` | Epidemiological datasets for validation | 🔬 Development |
| **Bayesian Analysis** | | | |
| | `tidybayes` | Bayesian posterior visualization | 📊 Vignettes |
| **Documentation & Tables** | | | |
| | `gt` | Publication-quality tables | 📊 Vignettes |
| | `flextable` | Flexible table formatting | 📊 Vignettes |
| | `broom` | Model output tidying | 📊 Vignettes |
</details>

## Contributing

Contributions to `mitey` are welcome! Please feel free to submit a pull request or open an [issue](https://github.com/kylieainslie/mitey/issues) to discuss potential improvements or report bugs.

<!--You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. -->
