---
title: "The Epidemiology of Scabies"
output: rmarkdown::html_vignette
bibliography: references.bib
csl: nature.csl
vignette: >
  %\VignetteIndexEntry{epidemiology_of_scabies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7, 
  fig.height=4
)

options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup, echo=FALSE, include = FALSE, results='hide'}
library(dplyr)
library(ggplot2)
library(brms)
library(devtools)
load_all()
```

## Introduction
Scabies is a neglected tropical disease caused by infestation of the skin with a microscopic mite (Sarcoptes scabiei). Symptoms are characterized by itchiness and rash at the site of infestation. Scabies affects around 400 million people per year, and accounts for a large proportion of skin disease in many low- and middle-income countries [@who]. A rise in scabies cases has been observed through out Europe in recent years [@deursen2022; @reichert2021; @redondo-bravo2021; @lugovic-mihic2020; @amato2019; @dona2023], putting pressure on local health services. Despite the considerable burden scabies poses annually, little is known about the disease dynamics of scabies transmission, such as the generation time (time between infection of an index case and a secondary case), serial interval (the time from onset of symptoms in an index case to the time of symptom onset in a secondary case), growth rate, and reproduction number (the average number of secondary cases resulting from one index case). A 1944 study by Mellanby [@mellanby1944] forms much of the basis of our current understanding of scabies transmission; however, the study was carried out on a small number of healthy volunteers which, and may not be representative of the natural history of scabies in populations in which scabies in widespread. 

In this work, we aim to estimate epidemiological parameters of scabies. It is critical to better understand the underlying disease dynamics of scabies to 1) assess current spread and 2) inform infection control policy. Some modelling work has been performed to study the potential impacts of intervention strategies, such as mass drug administration, on scabies transmission [@kinyanjui2018; @lydeamore2019; @tellioglu2023]; however, the values used to parameterise these models are based largely on Mellanby's 1944 study and use information about the mite life-cycle to approximate quantities such as latent period and infectious period. However, this information doesn't necessarily provide good estimates for transmission potential in reality. For example, Kinyanjui et al. [@kinyanjui2018] assume a latent period of 7-14 days to allow for the time for the time it takes for a fertilised female mite to reproduce and incorporate it into a susceptible-exposed-infectious model of scabies transmission. However, this assumes an equal chance of infectiousness despite the number of mites inhabiting an infested indivual. Mellanby [@mellanby1944] shows that the probability of onward transmission of scabies is rare when the number of mites on an individual is low and increases as the mite population grows. A better approximation of the disease process can be obtained using the generation time or serial interval; however, no such estimates exist for scabies. 

Additionally, the growth rate and reproduction number are quantities that describe how fast an infectious disease spreads. They can are also used to determine appropriate control measures to reduce the number of secondary infections so that an epidemic dies out. During the COVID-19 pandemic, many analyses were performed to determine what interventions were needed to "flatten the curve" whereby, the reproduction number (and growth rate) were reduced below one. However, in the case of scabies, the basic reproduction number has never been described. This makes it immensely difficult to determine what level of control measures are needed to contain disease spread. As evidence continues to suggest that scabies is a growing problem, it is important to describe the epidemiological characteristics of scabies.

In this study, we use epidemic curves of scabies outbreaks from the literature and estimate the serial interval. 

reproduction number of scabies.

## Data Sources

  - Kaburi et al. 2019 BMC Public Health https://doi.org/10.1186/s12889-019-7085-6
    - This paper describes an outbreak of scabies in a preschool in Ghana
  - Ariza et al.
    - this data is from a pre-school in Germany
  - Akunzirwe et al 2023. https://uniph.go.ug/an-outbreak-of-scabies-in-a-fishing-community-in-hoima-district-uganda-february%E2%88%92june-2022/
    - outbreak in a fishing community in Uganda in 2022. Data are provided weekly, so all infections for each week are attributed to the first day of the week
  - SCABIES OUTBREAK AMONG RESIDENTS OF A LONG TERM CARE FACILITY, CLARK COUNTY, NEVADA, 2015  file:///Users/kylieainslie/Downloads/Scabies%20%E2%80%93%20Clark%20[v%202015%20i%2021%20e%201.0]_BP.pdf
  - Tjon-Kon-Fat et al. (2021) Short report: The potential of PCR on skin flakes from bed linens for diagnosis of scabies in an outbreak. PLoS Negl Trop Dis 15(6): e0009485. https://doi.org/10.1371/journal.pntd.0009485
    - cases identified by week, date of onset is attributed to first day of the week
  - Larrosa A., et al. Nosocomial outbreak of scabies in a hospital in Spain. Euro Surveill. 2003;8(10):pii=429. https://doi.org/10.2807/esm.08.10.00429-en
    - Spain outbreak in a hospital 
    
## Serial Interval Analysis
Determining serial interval of scabies using epidemic curves from different sources.

We will calculate the index case-to-case (ICC) interval for each person by class the person with the greatest value for number of days since symptom onset will be considered the index case. The rest of the class members will have an ICC interval calculated as the number of days between their symptom onset and the index case. All analyses were performed in R 4.4.0 @R-base.

### Estimate serial interval distributions
Plot of the distribution of each study's serial interval. We assume a normal distribution.
```{r si_dist_plot, echo=TRUE, eval=TRUE}
# Plot serial interval curves
# Data preparation
x <- seq(0, 250, by = 1)
y1 <- dnorm(x, mean = 167.34442, sd = 9.71763)
y2 <- dnorm(x, mean = 98.4, sd = 8.542332)
y3 <- dnorm(x, mean = 122.92385, sd = 26.92035)
y4 <- dnorm(x, mean = 21.91776, sd = 15.23666)
y5 <- dnorm(x, mean = 110.71571, sd = 16.13879)
y6 <- dnorm(x, mean = 16.106246, sd = 2.421762)
# Create a data frame
my_data <- data.frame(
  x = rep(x, 6),
  y = c(y1, y2, y3, y4, y5, y6),
  group = factor(rep(c("Ghana outbreak", "Germany outbreak", "Uganda outbreak",
                       "Nevada (US) outbreak", "Netherlands outbreak", "Spain outbreak"), each = length(x)),
                 levels = c("Ghana outbreak", "Germany outbreak", "Uganda outbreak",
                            "Nevada (US) outbreak", "Netherlands outbreak", "Spain outbreak"))
)

# Create the plot
ggplot(my_data, aes(x = x, y = y, color = group)) +
  geom_line(linewidth = 1) +
  labs(title = "Estimated Serial Interval Distributions",
       x = "Time (days)",
       y = "Density") +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  )
```

### Meta-analysis
```{r, meta-analysis, eval = FALSE}
df_effect_sizes <- data.frame(
  mean_si = c(167.34442, 98.4, 122.92385, 21.91776, 110.71571, 16.106246),
  sd_si = c(9.71763, 8.542332, 26.92035, 15.23666, 16.13879, 2.421762),
  n = c(nrow(ghana_df), nrow(germany_df), nrow(uganda_long_df), nrow(nevada_long_df),
        nrow(dutch_long_df), nrow(spain_long_df)),
  country = c("Ghana", "Germany", "Uganda", "Nevada (USA)", "Netherlands", "Spain")
) %>%
  mutate(se_si = sd_si/sqrt(n))

# we will perform a Bayesian meta-analysis using the {brms} package
# specify priors
priors <- c(prior(normal(100,50), class = Intercept),
            prior(cauchy(0,1), class = sd))

# fit a random effects model
# convert sd to se
m.brm <- brm(mean_si|se(se_si) ~ 1 + (1|country),
             data = df_effect_sizes,
             prior = priors,
             iter = 5000,
             warmup = 2000,
             control = list(adapt_delta = 0.99, max_treedepth = 15))

summary(m.brm)
```
From `summary(m.brm)` we can see that the pooled mean serial interval is 95.57 (95% CI: 65.27, 124.83) and the pooled standard deviation is 15.17. In the output, `sd(Intercept)` represents the standard deviation of the random intercepts (the between-study variability). Since the value of the standard deviation of the random intercepts is large (58.31), this indicates that there is substantial heterogeneity among the studies. This is not surprising when looking at the estimated distributions of each study's serial interval shown in Figure X.

Create a forest plot with each study's estimate of mean SI and the pooled estimate.
```{r si_forest_plot, eval = FALSE}
forest_data <- df_effect_sizes %>%
  mutate(
    lower = mean_si - 1.96 * se_si,
    upper = mean_si + 1.96 * se_si,
    type = "Study"
  ) %>%
  select(country, mean_si, lower, upper, type) %>%
  rename(estimate = mean_si) %>%
  bind_rows(
    data.frame(
      country = "Pooled Estimate",
      estimate = posterior_summary(m.brm, variable = "Intercept")[, "Estimate"],
      lower = posterior_summary(m.brm, variable = "Intercept")[, "Q2.5"],
      upper = posterior_summary(m.brm, variable = "Intercept")[, "Q97.5"],
      type = "Pooled"
    )
  ) %>%
  # Ensure the pooled estimate is positioned last
  mutate(country = factor(country, levels = c(setdiff(country, "Pooled Estimate"), "Pooled Estimate")))

# Create the forest plot using ggplot2
ggplot(forest_data, aes(x = estimate, y = country, xmin = lower, xmax = upper, color = type)) +
  geom_point() +
  geom_errorbarh(height = 0.2) +
  geom_vline(xintercept = posterior_summary(m.brm, variable = "Intercept")[, "Estimate"], linetype = "dashed", color = "red") +
  labs(x = "Effect Size", y = "") +
  theme_minimal() +
  scale_color_manual(values = c("Study" = "blue", "Pooled" = "red")) +
  theme(legend.position = "none")
```

## Reproduction Number and Growth Rate

### Growth Rate
Using an exponential growth model and data from [NIVEL](https://www.nivel.nl/nl/zorg-en-ziekte-in-cijfers/actuele-cijfers-ziekten-per-week), the annual growth rate is r = 0.190.

[insert figure]

Determine time-varying reproduction number and characterise how it changes over time. We will use data from [NIVEL](https://www.nivel.nl/en) on the number of reported cases per week from 2011 to 2023 in the Netherlands. Data courtesy of Mariette Hooiveld.

# References

<div id="refs"></div>

# Appendix
TODO:
 - add text for meta-analysis
 - add text for reproduction number estimation
 - re-do growth rate and projection plot in ggplot and make into multi-panel figure
 
## Sensitivity Analyses
TODO:
  - add gamma distribution to Vink et al. method and re-run estimation
  - re-run meta-analysis assuming gamma distribution
  - potentially remove Uganda estimate from meta-analysis because it's community transmission and not a confined setting (e.g., school, hospital, care home)
  - check prior assumptions for meta-analysis

