---
title: "Flexible Number of Transmission Routes in Serial Interval Estimation"
author: "[Your name]"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Flexible Number of Transmission Routes in Serial Interval Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  fig.align  = "center",
  warning    = FALSE,
  message    = FALSE
)
```

```{r packages, echo = FALSE}
library(mitey)
library(ggplot2)
library(dplyr)
library(patchwork)   # to assemble multiple plots
library(fdrtool) 
```

```````````````{r theme, echo = FALSE}
COL_PRIMARY   <- "#2C6FAC"
COL_SECONDARY <- "#E05A2B"
COL_COPRIM    <- "#6BAE75"
COL_QUAT      <- "#9B59B6"
COL_ROUTE5    <- "#E8A838"
COL_ROUTE6    <- "#C0392B"
COL_ASYMP     <- "#BDBDBD"
COL_TRUE_SI   <- "#E74C3C"
```````````````

------------------------------------------------------------------------

# Introduction — Why does the serial interval matter?

The **serial interval** (SI) is defined as the time elapsed between symptom onset in a primary case (the infector) and symptom onset in the secondary case they directly infected [@vink2014].

<!-- FIGURE 1: schematic timeline infector → infectee with the SI indicated -->

```{=html}
<!-- [TO INSERT: timeline diagram showing two individuals, a transmission arrow,
     and a brace marking the SI between the two symptom onset dates] -->
```

```{r fig1, echo = FALSE, fig.cap = "Figure 1 — Definition of the serial interval"}

fig1_data <- data.frame(
  person = c("Infector", "Infector", "Infectee", "Infectee"),
  
  t_symp = c(0, 3, 8, 11),   # dates 
  y1      = c(1, 1, 1, 1), # position verticale
  y2      = c(1.1, 1.25, 0.75, 0.9 ),
  y3      = c(1.2, 1.2, 0.8, 0.8), 
  events = c("Infection of person 1","Symptom onset of person 1","Infection of person 2","Symptom onset of person 2")
)
 
# Repères de temps clés
t_symp_A <- 3
t_symp_B <- 11
 
fig1 <- ggplot() +
  # Lignes horizontales (timeline)
  geom_segment(
    data = fig1_data,
    aes(x = 0, xend = 18, y = y1, yend = y1),
    color = "grey70", linewidth = 0.6
  ) +
  # Points : symptom onset
  geom_point(
    data = fig1_data,
    aes(x = t_symp, y = y1, color = person),
    size = 5, show.legend = FALSE
  ) +
  # Étiquettes des personnes
  geom_text(
    data = fig1_data,
    aes(x = -0.5, y = y3, label = person, color = person),
    hjust = 1, fontface = "bold", size = 4, show.legend = FALSE
  ) +
  # Étiquettes "Symptom onset"
  geom_text(
    data = fig1_data,
    aes(x = t_symp, y = y2, label = events),
    size = 3, color = "grey30", lineheight = 0.85
  ) +
  # Accolade / flèche double pour l'IS
  geom_segment(
    aes(x = t_symp_A, xend = t_symp_B, y = 1.0, yend = 1.0),
    arrow = arrow(ends = "both", length = unit(0.2, "cm"), type = "closed"),
    color = COL_ROUTE5, linewidth = 1.2
  ) +
  annotate(
    "text", x = (t_symp_A + t_symp_B) / 2, y = 1.10,
    label = "Serial Interval (SI)", color = COL_ROUTE5,
    fontface = "bold", size = 4
  ) +
  # Ligne verticale pointillée pour chaque onset
  geom_vline(xintercept = c(t_symp_A, t_symp_B),
             linetype = "dashed", color = "grey60", linewidth = 0.5) +
  scale_color_manual(values = c("Infector" = COL_PRIMARY,
                                "Infectee" = COL_SECONDARY)) +
  scale_x_continuous(
    breaks = c(t_symp_A, t_symp_B),
    labels = c(paste0("Day ", t_symp_A), paste0("Day ", t_symp_B)),
    limits = c(-2, 19)
  ) +
  scale_y_continuous(limits = c(0.1, 1.9)) +
  labs(
    title   = "Figure 1 — Definition of the serial interval",
    x       = "Time (days)",
    y       = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y      = element_blank(),
    axis.ticks.y     = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title       = element_text(face = "bold", size = 13)
  )
 
fig1
 

```


In epidemiology, the serial interval is a fundamental parameter for several reasons:

-   **Quantifying the speed of spread.** A short SI means that successive generations of cases follow one another rapidly, making the epidemic harder to contain. A long SI leaves more time to detect and isolate cases before they transmit further.
-   **Estimating the reproduction number.** Both the basic reproduction number $R_0$ and the time-varying reproduction number $R_t$ are linked to the SI through the renewal equation: without an accurate SI estimate, any estimate of $R_t$ will be biased.
-   **Calibrating transmission models.** Compartmental models (SIR, SEIR, etc.) and renewal equation models use the SI as a direct input to simulate epidemic dynamics and evaluate the impact of interventions.
-   **Designing control measures.** The duration of quarantine, the contact-tracing window, and the timing of vaccination campaigns are all partly calibrated on the SI.

In practice, the SI is often confused with the **generation interval** (the time between infection of the primary case and infection of the secondary case). The two coincide when infectiousness begins at symptom onset, but they can diverge for pathogens with substantial pre-symptomatic transmission [@nishiura2010].

------------------------------------------------------------------------

# How is the serial interval estimated?

## Data requirements

Ideally, one would have transmission pairs (infector, infectee) identified with certainty through epidemiological investigation or genomic sequencing. In that setting, the SI is estimated directly as the difference in symptom onset dates.

In practice, such data are rare. The most common alternatives are:

| Approach | Data required | Main sources of bias |
|------------------------|------------------------|------------------------|
| Direct infector–infectee pairs | Contact-tracing surveys | Epidemic growth bias, right-truncation |
| Household data (ICC intervals) | Symptom onset dates within a household | Multiple transmission routes (see §3) |
| Bayesian reconstruction | Partial transmission tree | Strongly prior-dependent |

## Classical biases and corrections

Several biases affect SI estimation from observed data:

-   **Epidemic growth bias.** During epidemic growth, infector–infectee pairs with short SIs are over-represented in surveillance data, because recent infectors are more likely to still be under observation [@park2019].
-   **Right-truncation.** Secondary cases whose symptoms appear after the study end date are not observed, artificially shortening the estimated SI.
-   **Negative serial intervals.** When the secondary case develops symptoms before the identified infector (pre-symptomatic transmission), the observed SI is negative.

These corrections are well documented in the literature [@ali2023] but require additional assumptions. It is precisely to sidestep some of these difficulties that household-based ICC methods were developed.

------------------------------------------------------------------------

# The Vink et al. method

## ICC data (*Index Case-to-Case intervals*)

In a household or institutional setting (care home, school classroom), the **index case** — the first person to show symptoms — is identified, and the symptom onset date of every other case is recorded. The difference between the index case onset date and each subsequent case onset date is an **ICC interval**.

```{=html}
<!-- FIGURE 2: schematic of a household with an index case and several secondary
     cases, showing the different ICC values -->
```

```{r fig2, echo = FALSE}
household <- data.frame(
  person = c("Index case", "Case B", "Case C", "Case D"),
  t_symp = c(0, 8, 15, 22),
  y      = c(2.0, 1.5, 1.0, 0.5),
  role   = c("index", "secondary", "secondary", "secondary")
)
 
fig2 <- ggplot(household) +
  # Timeline commune (axe horizontal)
  geom_segment(aes(x = -1, xend = 26, y = y, yend = y),
               color = "grey80", linewidth = 0.5) +
  # Points symptom onset
  geom_point(aes(x = t_symp, y = y, color = role), size = 5,
             show.legend = FALSE) +
  # Étiquettes personnes
  geom_text(aes(x = -1.2, y = y, label = person, color = role),
            hjust = 1, fontface = "bold", size = 3.8, show.legend = FALSE) +
  # Flèches ICC (du cas index vers chaque autre cas)
  geom_segment(
    data = household %>% filter(role == "secondary"),
    aes(x = 0, xend = t_symp, y = y, yend = y),
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = COL_SECONDARY, linewidth = 1, linetype = "solid"
  ) +
  # Étiquettes ICC
  geom_label(
    data = household %>% filter(role == "secondary"),
    aes(x = t_symp / 2, y = y + 0.18,
        label = paste0("ICC = ", t_symp, " days")),
    size = 3, color = COL_SECONDARY, fill = "white",
    label.size = 0.2, label.padding = unit(0.15, "lines")
  ) +
  # Ligne verticale cas index
  geom_vline(xintercept = 0, linetype = "dashed",
             color = COL_PRIMARY, linewidth = 0.7) +
  annotate("text", x = 0.5, y = 2.35, label = "Index case\nonset",
           color = COL_PRIMARY, size = 3.2, hjust = 0, lineheight = 0.85) +
  scale_color_manual(values = c("index"     = COL_PRIMARY,
                                "secondary" = COL_SECONDARY)) +
  scale_x_continuous(limits = c(-5.5, 27),
                     breaks = c(0, 8, 15, 22),
                     labels = c("Day 0", "Day 8", "Day 15", "Day 22")) +
  scale_y_continuous(limits = c(0.2, 2.5)) +
  labs(
    title = "Figure 2 — ICC intervals within a household",
    x     = "Time (days)",
    y     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold", size = 13)
  )
 
fig2
```



```{=html}
<!-- [TO INSERT: timeline of a household, index case at t = 0, arrows pointing
     to cases at t = ICC_1, ICC_2, etc.] -->
```

The appeal of ICC data is their availability: they do not require identifying exactly who infected whom, only that symptom onset dates are recorded within a well-defined group.

```         



## The mixture model: transmission routes

The key insight of Vink et al. is that each observed ICC can correspond to **several distinct transmission scenarios**, called routes:

-   **Co-primary (CP):** Both the index case and the observed case were infected by the same external source at roughly the same time. The ICC reflects variability in incubation, not the SI. Expected distribution: half-normal centred at zero.
-   **Primary-Secondary (PS):** The index case directly infected the observed case. ICC ≈ SI. Distribution: $\mathcal{N}(\mu, \sigma^2)$.
-   **Primary-Tertiary (PT):** The index case infected an asymptomatic intermediate, who then infected the observed case. ICC ≈ 2 × SI. Distribution: $\mathcal{N}(2\mu, 2\sigma^2)$.
-   **Primary-Quaternary (PQ):** Two asymptomatic intermediates between the index case and the observed case. ICC ≈ 3 × SI. Distribution: $\mathcal{N}(3\mu, 3\sigma^2)$.

```{=html}
<!-- FIGURE 3: diagram of the 4 transmission routes, with the transmission chain
     for each route and the expected ICC distribution -->
```

```{r fig3}

```


```{=html}
<!-- [TO INSERT: 4 panels, one per route, showing the transmission chain
     (nodes with or without symptoms) and the ICC probability density] -->
```

The observed ICC density is therefore a **mixture** of these components:

$$f_{\text{ICC}}(x) = w_{\text{CP}} \cdot f_{\text{CP}}(x) +
                       w_{\text{PS}} \cdot f_{\text{PS}}(x) +
                       w_{\text{PT}} \cdot f_{\text{PT}}(x) +
                       w_{\text{PQ}} \cdot f_{\text{PQ}}(x)$$

where the $w_k$ are the route weights (with $\sum w_k = 1$), and $\mu$, $\sigma$ are the underlying SI parameters to be estimated.

## The EM algorithm

The parameters $(\mu, \sigma, w_{\text{CP}}, w_{\text{PS}}, w_{\text{PT}},
w_{\text{PQ}})$ are estimated using an **Expectation-Maximisation (EM)** algorithm:

1.  **E-step.** For each observation $x_i$, compute the posterior probability that it belongs to each transmission route.
2.  **M-step.** Update $\mu$ and $\sigma$ by maximising the weighted likelihood, and recompute the weights $w_k$.

These two steps are repeated until convergence. For further technical details, refer to the *Quick Start Guide* vignette of the `mitey` package and to @vink2014.

```{r basic-example, eval=FALSE}
# Basic estimation with 4 routes (default behaviour)
result <- si_estim(icc_data, dist = "normal", n_routes = 4)
result$mean   # Estimated mean serial interval (days)
result$sd     # Estimated standard deviation (days)
```

------------------------------------------------------------------------

# The key limitation: an arbitrarily fixed number of routes

## The problem

In the original formulation of Vink et al., the number of transmission routes is **fixed at four** (CP, PS, PT, PQ). This choice is reasonable for many common infectious diseases, where it is unlikely to observe more than two or three asymptomatic intermediates between the index case and the observed case.

However, this choice is **arbitrary** and can be ill-suited in two contrasting situations.

## Highly asymptomatic diseases: the risk of underestimating the number of routes

For some diseases, a substantial fraction of infections remains asymptomatic. A notable example is **scabies** (*Sarcoptes scabiei*), for which the incubation period can last several weeks, and infested individuals can be contagious without showing any visible symptoms for a long time.

In this context, long transmission chains are plausible: the index case may have infected one asymptomatic intermediate, who infected another, who infected yet another, before the observed case finally shows symptoms.

```{=html}
<!-- FIGURE 4: schematic of a long transmission chain for a highly asymptomatic
     disease, showing 5 or 6 links between the index case and the observed case -->
```

```{=html}
<!-- [TO INSERT: linear chain of nodes, some greyed out (asymptomatic),
     with the total ICC indicated] -->
```

If such long chains are frequent but the algorithm is limited to 4 routes, the PT or PQ component will absorb ICC values that actually correspond to routes of order 5 or 6. This leads to **overestimation of the SI**, because the algorithm assigns an ICC of $5\mu$ to the component centred at $3\mu$.

```{=html}
<!-- FIGURE 5: comparison of the estimated vs. true density for different numbers
     of routes (simulation with n_routes = 4 vs. n_routes = 6) -->
```

```{=html}
<!-- [TO INSERT: two density curves overlaid on the same histogram — one with
     4 routes (shifted right, overestimation), one with 6 routes (correct fit),
     and a vertical line at the true SI] -->
```

## Highly symptomatic diseases: the risk of over-parameterisation

Conversely, for a disease where almost all infections are symptomatic (e.g. measles or smallpox), it is unlikely that several asymptomatic intermediates follow one another in a single transmission chain. In this case, **modelling 4 routes introduces unnecessary parameters** (the weights for PT and PQ routes will be estimated near zero), at the cost of precision and numerical stability of the EM algorithm.

A model with only 2 or 3 routes would be more parsimonious and would yield more stable estimates.

## Summary

```{r routes-table, echo=FALSE}
df_routes <- data.frame(
  Context = c(
    "Highly asymptomatic disease",
    "Moderately asymptomatic disease",
    "Highly symptomatic disease"
  ),
  Examples = c(
    "Scabies, some COVID-19 variants",
    "Influenza, SARS-CoV-2 (Delta variant)",
    "Measles, smallpox"
  ),
  `Recommended routes` = c(
    "\u2265 5 \u2013 8",
    "3 \u2013 5 (default = 4)",
    "2 \u2013 3"
  ),
  `Risk if n_routes too low` = c(
    "Overestimation of the SI",
    "Low",
    "\u2013"
  ),
  `Risk if n_routes too high` = c(
    "\u2013",
    "Low",
    "Over-parameterisation, instability"
  )
)
knitr::kable(
  df_routes,
  caption = "Impact of the number of routes depending on the epidemiological profile of the disease",
  col.names = c("Context", "Examples", "Recommended routes",
                "Risk if too low", "Risk if too high")
)
```

------------------------------------------------------------------------

# The new feature: `n_routes` as a user parameter

## Principle

The central modification brought to `mitey` is to **expose the number of transmission routes as an explicit parameter** of `si_estim()`, rather than hard-coding it to 4. The user can now specify any integer $\geq 2$.

The default value remains `n_routes = 4` to ensure backward compatibility with the original behaviour.

## What changed in the code

The change affects several functions in a consistent way:

| Function | Role | Modification |
|------------------------|------------------------|------------------------|
| `si_estim()` | Main estimation function | `n_routes` parameter added |
| `f_norm()` | Mixture density (normal distribution) | Adapted for $n$ components |
| `f_gam()` | Mixture density (gamma distribution) | Adapted for $n$ components |
| `integrate_component()` | Numerical integration per component | Unchanged (already generic) |
| `integrate_components_wrapper()` | Integration across all components | `n_routes` parameter added |
| `plot_si_fit()` | Fitted distribution visualisation | Adapted for $n$ components |
| `plot_si_fit_result()` | Visualisation wrapper | `n_routes` forwarded |

The component numbering logic is as follows (normal distribution):

-   Component 1: Co-Primary route (half-normal)
-   Components $2i$ and $2i+1$ for $i = 1, \ldots, n_{\text{routes}} - 1$: positive and negative routes for transmission route $i+1$

For the gamma distribution (which only produces positive values), only even components are used, giving $n_{\text{routes}}$ components in total.

## Updated function signatures

```{r signatures, eval=FALSE}
# Estimation with a custom number of routes
result <- si_estim(
  dat      = icc_data,
  dist     = "normal",   # or "gamma"
  n_routes = 6L          # desired number of routes (default: 4)
)

# Direct visualisation of the result
plot_si_fit_result(result, dat = icc_data, dist = "normal")
```

The function returns the same fields as before, with two additions:

-   `n_routes`: the number of routes used (useful for traceability)
-   `loglik`: the log-likelihood of the fitted model (useful for model comparison, see §6)

------------------------------------------------------------------------

# Choosing the right number of routes: `compare_n_routes()`

## Model comparison principle

Increasing the number of routes necessarily improves the fit to the data (the log-likelihood can only increase or stay the same), but at the cost of **greater complexity** and a risk of overfitting. A **penalised criterion** is therefore needed to find the right balance.

The package implements two classical criteria:

$$\text{AIC} = -2\ell + 2p \qquad \text{BIC} = -2\ell + p \ln(n)$$

where $\ell$ is the log-likelihood, $p$ the number of free parameters, and $n$ the sample size. The BIC penalises complexity more heavily and therefore favours more parsimonious models — it is generally recommended for selecting the number of components in a mixture model.

The number of free parameters is: - **Normal distribution:** $p = 2 + (2 \times n_{\text{routes}} - 2)$ (2 for $\mu$ and $\sigma$, plus the free weights) - **Gamma distribution:** $p = 2 + (n_{\text{routes}} - 1)$

## The `compare_n_routes()` function

```{r compare-n-routes-example, eval=FALSE}
# Compare models from 2 to 8 routes
comparison <- compare_n_routes(
  dat          = icc_data,
  n_routes_max = 8L,
  dist         = "normal",
  n            = 50        # number of EM iterations per model
)
```

The function returns a summary table and automatically displays the fitted distribution plots for each value of `n_routes` tested:

```{=html}
<!-- FIGURE 6: AIC and BIC as a function of the number of routes, with the
     minimum highlighted -->
```

```{=html}
<!-- [TO INSERT: line plot of AIC and BIC vs. n_routes, minimum point
     highlighted, possibly with bar charts for each model] -->
```

```{r compare-n-routes-output, eval=FALSE}
# Structure of the output
# n_routes | mu   | sigma | loglik | aic    | bic    | n_params | converged | iterations
# 2        | ...  | ...   | ...    | ...    | ...    | ...      | TRUE      | ...
# 3        | ...  | ...   | ...    | ...    | ...    | ...      | TRUE      | ...
# ...
```

## Starting heuristics: `n_optimal1()`, `n_optimal2()`, `n_optimal3()`

Three simple data-driven heuristics are available to guide the initial choice of `n_routes_max`:

```{r heuristics, eval=FALSE}
# Heuristic 1: based on the 95th percentile and the mean
n_optimal1(icc_data)   # as.integer(quantile(dat, 0.95) / mean(dat))

# Heuristic 2: based on the coefficient of variation
n_optimal2(icc_data)   # as.integer(sd(dat) / mean(dat)) + 2

# Heuristic 3: based on the relative range
n_optimal3(icc_data)   # as.integer(max(dat) / sd(dat))
```

These heuristics are not rigorous estimators but reasonable **starting points** for exploring the `n_routes` grid. It is recommended to use them to set `n_routes_max` in `compare_n_routes()`, and then rely on AIC/BIC for the final selection.

------------------------------------------------------------------------

# Applied examples

## Example 1 — Highly asymptomatic disease (simulated data)

We simulate a disease with a high probability of asymptomatic infection. This results in ICC values that may correspond to long transmission chains: some observed ICCs are 5× or 6× the true SI.

```{r simulation-high-asymptomatic, eval=FALSE}
set.seed(42)

true_mu    <- 10   # True SI: 10 days
true_sigma <- 2    # True standard deviation: 2 days
N          <- 300

# Route weights: many long routes
weights_asym <- c(0.10,   # CP
                  0.20,   # PS
                  0.20,   # PT
                  0.20,   # PQ
                  0.15,   # Route 5
                  0.15)   # Route 6

# Simulate ICC intervals
# [TO COMPLETE following the same scheme as simulated_ICC_interval.R]

# Estimation with 4 routes (original method)
result_4 <- si_estim(icc_asym, dist = "normal", n_routes = 4)

# Estimation with the correct number of routes
result_6 <- si_estim(icc_asym, dist = "normal", n_routes = 6)

cat("True mean SI      :", true_mu, "days\n")
cat("Estimated (4 routes):", round(result_4$mean, 2), "days\n")
cat("Estimated (6 routes):", round(result_6$mean, 2), "days\n")
```

```{=html}
<!-- FIGURE 7: two histograms side by side (or overlaid) with the estimated
     density curve for n_routes = 4 (red) and n_routes = 6 (blue),
     and a vertical line indicating the true SI -->
```

<!-- [TO INSERT: plot_si_fit_result() for both models, assembled with patchwork] -->

```{r compare-routes-asym, eval=FALSE}
# Formal comparison by AIC/BIC
comparison_asym <- compare_n_routes(
  dat          = icc_asym,
  n_routes_max = 8L,
  dist         = "normal"
)
```

```{=html}
<!-- FIGURE 8: AIC/BIC vs. n_routes for the asymptomatic disease,
     showing that the minimum is at n_routes > 4 -->
```

<!-- [TO INSERT] -->

## Example 2 — Highly symptomatic disease (simulated data)

We now simulate a disease where almost all cases are symptomatic: long routes (PT, PQ) are very rare.

```{r simulation-high-symptomatic, eval=FALSE}
set.seed(42)

true_mu    <- 5    # True SI: 5 days
true_sigma <- 1.5

# Route weights: almost exclusively CP and PS
weights_symp <- c(0.15,   # CP
                  0.80,   # PS
                  0.05)   # PT (route 3, very rare)

# Simulate ICC intervals
# [TO COMPLETE]

# Formal comparison
comparison_symp <- compare_n_routes(
  dat          = icc_symp,
  n_routes_max = 6L,
  dist         = "normal"
)
```

```{=html}
<!-- FIGURE 9: AIC/BIC vs. n_routes for the symptomatic disease,
     showing that the minimum is at n_routes = 2 or 3 -->
```

<!-- [TO INSERT] -->

## Example 3 — Real data: scabies

Scabies is a parasitic skin disease caused by the mite *Sarcoptes scabiei*, characterised by a long incubation period (up to 6 weeks for a primary infestation) and transmission that can go unnoticed for several weeks. It therefore represents a textbook case for the benefit of allowing a higher number of routes.

```{r scabies-example, eval=FALSE}
# Load scabies data (published studies)
# file_path <- here("vignettes", "data", "si_data.rds")
# scabies_data <- readRDS(file_path)

# Select one study
# icc_scabies <- scabies_data %>%
#   filter(study == "Akunzirwe et al.") %>%
#   pull(icc_interval)

# Compare the optimal number of routes
comparison_scabies <- compare_n_routes(
  dat          = icc_scabies,
  n_routes_max = 8L,
  dist         = "normal"
)
```

```{=html}
<!-- FIGURE 10: output of compare_n_routes() on the scabies data,
     histogram with the fitted density for the BIC-selected optimal n_routes -->
```

<!-- [TO INSERT] -->

------------------------------------------------------------------------

# Conclusion and practical recommendations

## Summary

This vignette has presented the extension brought to the `si_estim()` function of the `mitey` package, which now allows the user to freely specify the number of transmission routes $n_{\text{routes}}$.

The key takeaways are:

-   The number of routes is a **modelling choice** that should be guided by the biology of the disease and validated by statistical criteria (AIC/BIC).
-   Using **too few routes** for a highly asymptomatic disease leads to overestimation of the SI.
-   Using **too many routes** for a highly symptomatic disease introduces unnecessary parameters and can destabilise the EM algorithm.
-   The `compare_n_routes()` function automates the comparison and allows the optimal model to be selected objectively.

## Recommended workflow

```{r recommended-workflow, eval=FALSE}
# Step 1: get a ballpark for the number of routes using the heuristics
n1 <- n_optimal1(icc_data)
n2 <- n_optimal2(icc_data)
n3 <- n_optimal3(icc_data)

cat("Heuristics:", n1, n2, n3, "\n")
n_max_explore <- max(n1, n2, n3) + 2   # safety margin

# Step 2: compare models from 2 to n_max_explore routes
comparison <- compare_n_routes(
  dat          = icc_data,
  n_routes_max = n_max_explore,
  dist         = "normal"
)

# Step 3: estimate with the BIC-optimal number of routes
n_opt <- comparison$n_routes[which.min(comparison$bic)]
result_final <- si_estim(icc_data, dist = "normal", n_routes = n_opt)

# Step 4: visualise the final fit
plot_si_fit_result(result_final, dat = icc_data, dist = "normal")
```

## Limitations and perspectives

-   The method still relies on the assumption that the index case is indeed the person with the earliest symptom onset date in the household, which may not always hold.
-   The maximum number of routes is limited in practice by sample size: with small datasets, the weights of high-order routes will be estimated with large uncertainty. As a rule of thumb, avoid exploring routes whose expected ICC ($k \times \mu$) exceeds the 95th percentile of the observed data.
-   A natural extension would be a fully Bayesian formulation of the mixture model, which would provide uncertainty intervals on all parameters jointly and handle small samples more gracefully.

------------------------------------------------------------------------

# References

::: {#refs}
:::

------------------------------------------------------------------------

*This document was generated with `rmarkdownr packageVersion("rmarkdown")` and `miteyr packageVersion("mitey")`.*
