---
title: "Flexible Number of Transmission Routes and accounting for date uncertainty in Serial Interval Estimation"
author: "Simon Carreel"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Flexible Number of Transmission Routes in Serial Interval Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  fig.align  = "center",
  warning    = TRUE,
  message    = FALSE
)
```

```{r packages, echo = FALSE}
library(mitey)
library(ggplot2)
library(dplyr)
library(patchwork)   
library(fdrtool) 
```

```{r theme, echo = FALSE}
COL_PRIMARY   <- "#2C6FAC"
COL_SECONDARY <- "#E05A2B"
COL_COPRIM    <- "#6BAE75"
COL_QUAT      <- "#9B59B6"
COL_ROUTE5    <- "#E8A838"
COL_ROUTE6    <- "#C0392B"
COL_ASYMP     <- "#BDBDBD"
COL_TRUE_SI   <- "#E74C3C"
```

------------------------------------------------------------------------

# Introduction --- Why does the serial interval matter?

The **serial interval** (SI) is defined as the time elapsed between symptom onset in a primary case (the infector) and symptom onset in the secondary case they directly infected [@vink2014].

```{r SI_definition_figure_1, echo = FALSE, fig.cap = "Figure 1 — Definition of the serial interval"}

fig1_data <- data.frame(
  person = c("Infector", "Infectee"),
  t_inf  = c(0, 6),    # dates d'infection (non observées)
  t_symp = c(4, 14),   # dates d'apparition des symptômes
  y      = c(1.5, 0.5) # position verticale
)

# Repères de temps clés
t_symp_A <- 4
t_symp_B <- 14

fig1 <- ggplot() +
  # Lignes horizontales (timeline)
  geom_segment(
    data = fig1_data,
    aes(x = 0, xend = 18, y = y, yend = y),
    color = "grey70", linewidth = 0.6
  ) +
  # Points : symptom onset
  geom_point(
    data = fig1_data,
    aes(x = t_symp, y = y, color = person),
    size = 5, show.legend = FALSE
  ) +
  # Étiquettes des personnes
  geom_text(
    data = fig1_data,
    aes(x = -0.5, y = y, label = person, color = person),
    hjust = 1, fontface = "bold", size = 4, show.legend = FALSE
  ) +
  # Étiquettes "Symptom onset"
  geom_text(
    data = fig1_data,
    aes(x = t_symp, y = y + 0.18, label = "Symptom\nonset"),
    size = 3, color = "grey30", lineheight = 0.85
  ) +
  # Accolade / flèche double pour l'IS
  geom_segment(
    aes(x = t_symp_A, xend = t_symp_B, y = 1.0, yend = 1.0),
    arrow = arrow(ends = "both", length = unit(0.2, "cm"), type = "closed"),
    color = COL_ROUTE5, linewidth = 1.2
  ) +
  annotate(
    "text", x = (t_symp_A + t_symp_B) / 2, y = 1.10,
    label = "Serial Interval (SI)", color = COL_ROUTE5,
    fontface = "bold", size = 4
  ) +
  # Ligne verticale pointillée pour chaque onset
  geom_vline(xintercept = c(t_symp_A, t_symp_B),
             linetype = "dashed", color = "grey60", linewidth = 0.5) +
  scale_color_manual(values = c("Infector" = COL_PRIMARY,
                                "Infectee" = COL_SECONDARY)) +
  scale_x_continuous(
    breaks = c(t_symp_A, t_symp_B),
    labels = c(paste0("Day ", t_symp_A), paste0("Day ", t_symp_B)),
    limits = c(-2, 19)
  ) +
  scale_y_continuous(limits = c(0.1, 1.9)) +
  labs(
    title   = "Figure 1 — Definition of the serial interval",
    x       = "Time (days)",
    y       = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y      = element_blank(),
    axis.ticks.y     = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title       = element_text(face = "bold", size = 13)
  )

fig1
 

```

In epidemiology, the serial interval is a fundamental parameter for several reasons:

-   **Quantifying the speed of spread.** A short SI means that successive generations of cases follow one another rapidly, making the epidemic harder to contain. A long SI leaves more time to detect and isolate cases before they transmit further.
-   **Estimating the reproduction number.** Both the basic reproduction number $R_0$ and the time-varying reproduction number $R_t$ are linked to the SI through the renewal equation: without an accurate SI estimate, any estimate of $R_t$ will be biased.
-   **Calibrating transmission models.** Compartmental models (SIR, SEIR, etc.) and renewal equation models use the SI as a direct input to simulate epidemic dynamics and evaluate the impact of interventions.
-   **Designing control measures.** The duration of quarantine, the contact-tracing window, and the timing of vaccination campaigns are all partly calibrated on the SI.

In practice, the SI is often confused with the **generation interval** (the time between infection of the primary case and infection of the secondary case).

------------------------------------------------------------------------

# How is the serial interval estimated?

## Data requirements

Ideally, one would have transmission pairs (infector, infectee) identified with certainty through epidemiological investigation or genomic sequencing. In that setting, the SI is estimated directly as the difference in symptom onset dates.

In practice, such data are rare. The most common alternatives are:

| Approach                        | Data required                          | Main sources of bias                   |
|------------------------|------------------------|------------------------|
| Direct infector--infectee pairs | Contact-tracing surveys                | Epidemic growth bias, right-truncation |
| Outbreak data (ICC intervals)   | Symptom onset dates within a household | Multiple transmission routes (see §3)  |
| Bayesian reconstruction         | Partial transmission tree              | Strongly prior-dependent               |

## Classical biases and corrections

Several biases affect SI estimation from observed data:

-   **Epidemic growth bias.** During epidemic growth, infector--infectee pairs with short SIs are over-represented in surveillance data, because recent infectors are more likely to still be under observation.
-   **Right-truncation.** Secondary cases whose symptoms appear after the study end date are not observed, artificially shortening the estimated SI.
-   **Negative serial intervals.** When the secondary case develops symptoms before the identified infector (pre-symptomatic transmission), the observed SI is negative.

These corrections are well documented in the literature but require additional assumptions. It is precisely to sidestep some of these difficulties that household-based ICC methods were developed.

## Date reporting uncertainty

A less-discussed but practically important source of bias in serial interval estimation
concerns the **accuracy of the reported symptom onset date**. In many surveillance
systems, the date recorded in the database is not the true date of symptom onset, but
rather the date at which the case was *reported* to the health authorities.

Several mechanisms can introduce this discrepancy:

- **Weekly reporting cycles.** In some institutional settings (care homes, schools),
  health authorities collect information only once a week. A person who developed
  symptoms on a Wednesday may only be recorded on the following Monday, introducing
  an error of up to 6 days.
- **No weekend reporting.** Some surveillance systems do not register new cases on
  Saturdays and Sundays. Cases with symptom onset on Friday evening or Saturday will
  systematically be reported on the following Monday.
- **Grouped declaration.** Physicians or hospitals sometimes submit batches of cases
  at fixed intervals, leading to artificial clustering of reported onset dates.

The consequence for ICC interval estimation is that the observed ICC is not
$\Delta t = t_{\text{onset, secondary}} - t_{\text{onset, index}}$, but rather
$\Delta t^* = t_{\text{reported, secondary}} - t_{\text{reported, index}}$, which
can differ from the true value by up to $2 \times \texttt{wind}$ days (if both the
index and the secondary case are subject to reporting delays in opposite directions).

The `mitey` package addresses this through the `wind` parameter (reporting **wind**ow),
described in detail in Section 4.

------------------------------------------------------------------------

# The Vink et al. method

## ICC data (*Index Case-to-Case intervals*)

In a household or institutional setting (care home, school classroom), the **index case** --- the first person to show symptoms --- is identified, and the symptom onset date of every other case is recorded. The difference between the index case onset date and each subsequent case onset date is an **ICC interval**.

```{r ICC_interval_figure_2, echo = FALSE}
household <- data.frame(
  person = c("Index case", "Case B", "Case C", "Case D"),
  t_symp = c(0, 8, 15, 22),
  y      = c(2.0, 1.5, 1.0, 0.5),
  role   = c("index", "secondary", "secondary", "secondary")
)
 
fig2 <- ggplot(household) +
  # Timeline commune (axe horizontal)
  geom_segment(aes(x = -1, xend = 26, y = y, yend = y),
               color = "grey80", linewidth = 0.5) +
  # Points symptom onset
  geom_point(aes(x = t_symp, y = y, color = role), size = 5,
             show.legend = FALSE) +
  # Étiquettes personnes
  geom_text(aes(x = -1.2, y = y, label = person, color = role),
            hjust = 1, fontface = "bold", size = 3.8, show.legend = FALSE) +
  # Flèches ICC (du cas index vers chaque autre cas)
  geom_segment(
    data = household %>% filter(role == "secondary"),
    aes(x = 0, xend = t_symp, y = y, yend = y),
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = COL_SECONDARY, linewidth = 1, linetype = "solid"
  ) +
  # Étiquettes ICC
  geom_label(
    data = household %>% filter(role == "secondary"),
    aes(x = t_symp / 2, y = y + 0.18,
        label = paste0("ICC = ", t_symp, " days")),
    size = 3, color = COL_SECONDARY, fill = "white",
    label.size = 0.2, label.padding = unit(0.15, "lines")
  ) +
  # Ligne verticale cas index
  geom_vline(xintercept = 0, linetype = "dashed",
             color = COL_PRIMARY, linewidth = 0.7) +
  annotate("text", x = 0.5, y = 2.35, label = "Index case\nonset",
           color = COL_PRIMARY, size = 3.2, hjust = 0, lineheight = 0.85) +
  scale_color_manual(values = c("index"     = COL_PRIMARY,
                                "secondary" = COL_SECONDARY)) +
  scale_x_continuous(limits = c(-5.5, 27),
                     breaks = c(0, 8, 15, 22),
                     labels = c("Day 0", "Day 8", "Day 15", "Day 22")) +
  scale_y_continuous(limits = c(0.2, 2.5)) +
  labs(
    title = "Figure 2 — ICC intervals within a household",
    x     = "Time (days)",
    y     = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold", size = 13)
  )
 
fig2
```

The appeal of ICC data is their availability: they do not require identifying exactly who infected whom, only that symptom onset dates are recorded within a well-defined group.

## The mixture model: transmission routes

The key insight of Vink et al. is that each observed ICC can correspond to **several distinct transmission scenarios**, called routes:

-   **Co-primary (CP):** Both the index case and the observed case were infected by the same external source at roughly the same time. The ICC reflects variability in incubation, not the SI. Expected distribution: half-normal centred at zero.
-   **Primary-Secondary (PS):** The index case directly infected the observed case. ICC ≈ SI. Distribution: $\mathrm{N}(\mu, \sigma^2)$.
-   **Primary-Tertiary (PT):** The index case infected an asymptomatic intermediate, who then infected the observed case. ICC ≈ 2 × SI. Distribution: $\mathrm{N}(2\mu, 2\sigma^2)$.
-   **Primary-Quaternary (PQ):** Two asymptomatic intermediates between the index case and the observed case. ICC ≈ 3 × SI. Distribution: $\mathrm{N}(3\mu, 3\sigma^2)$.

```{r four_transmission_routes_fig3, echo = FALSE}
mu_ex    <- 10
sigma_ex <- 2
x_seq    <- seq(-5, 40, length.out = 400)

# Densités pour chaque route
dens_CP <- fdrtool::dhalfnorm(x_seq,
                              theta = sqrt(pi / 2) / (sqrt(2) * sigma_ex))
dens_PS <- dnorm(x_seq, mean = mu_ex,      sd = sigma_ex)
dens_PT <- dnorm(x_seq, mean = 2 * mu_ex,  sd = sqrt(2) * sigma_ex)
dens_PQ <- dnorm(x_seq, mean = 3 * mu_ex,  sd = sqrt(3) * sigma_ex)

dens_df <- data.frame(
  x    = rep(x_seq, 4),
  dens = c(dens_CP, dens_PS, dens_PT, dens_PQ),
  route = rep(
    c("Co-Primary (CP)\nICC ~ 0",
      "Primary-Secondary (PS)\nICC ~ SI",
      "Primary-Tertiary (PT)\nICC ~ 2 × SI",
      "Primary-Quaternary (PQ)\nICC ~ 3 × SI"),
    each = length(x_seq)
  )
)
dens_df$route <- factor(dens_df$route, levels = unique(dens_df$route))

route_colors <- c(
  "Co-Primary (CP)\nICC ~ 0"              = COL_COPRIM,
  "Primary-Secondary (PS)\nICC ~ SI"      = COL_PRIMARY,
  "Primary-Tertiary (PT)\nICC ~ 2 × SI"   = COL_SECONDARY,
  "Primary-Quaternary (PQ)\nICC ~ 3 × SI" = COL_QUAT
)

# Schémas de chaînes de transmission (coordonnées des nœuds)
make_chain <- function(n_asymp, y_base = 0.7, x_start = 0, x_step = mu_ex) {
  
  if (n_asymp == -1) {
    # Cas Co-Primary : seulement 2 nœuds, tous les deux symptomatiques
    x_nodes <- x_start + c(0, x_step)
    symp    <- c(TRUE, TRUE)
  } else {
    n_total <- n_asymp + 2  # index + intermédiaires + cas observé
    x_nodes <- x_start + (0:(n_total - 1)) * x_step
    symp    <- c(TRUE, rep(FALSE, n_asymp), TRUE)
  }
  
  data.frame(
    x    = x_nodes,
    y    = y_base,
    symp = symp
  )
}

make_arrows <- function(nodes) {
  data.frame(
    x    = head(nodes$x, -1),
    xend = tail(nodes$x, -1),
    y    = head(nodes$y, -1),
    yend = tail(nodes$y, -1)
  )
}

# Un panneau par route
make_route_panel <- function(route_label, n_asymp, color,
                             mu_val, sigma_val, x_seq) {
  nodes  <- make_chain(n_asymp, y_base = 0.72)
  arrows <- make_arrows(nodes)

  # Densité
  if (n_asymp == -1) {
    # CP : demi-normale
    dens <- fdrtool::dhalfnorm(x_seq,
                               theta = sqrt(pi / 2) / (sqrt(2) * sigma_val))
    mean_val <- NA
  } else {
    route_i  <- n_asymp + 1
    dens     <- dnorm(x_seq,
                      mean = route_i * mu_val,
                      sd   = sqrt(route_i) * sigma_val)
    mean_val <- route_i * mu_val
  }
  dens_panel <- data.frame(x = x_seq, y = dens)

  p <- ggplot() +
    # ---- chaîne de transmission ----
    geom_segment(
      data = arrows,
      aes(x = x, xend = xend, y = y, yend = yend),
      arrow = arrow(length = unit(0.15, "cm"), type = "closed"),
      color = "grey50", linewidth = 0.7
    ) +
    # nœuds symptomatiques (ronds pleins)
    geom_point(
      data = nodes %>% filter(symp),
      aes(x = x, y = y),
      color = color, size = 5
    ) +
    # nœuds asymptomatiques (ronds vides gris)
    geom_point(
      data = nodes %>% filter(!symp),
      aes(x = x, y = y),
      color = COL_ASYMP, size = 5, shape = 1, stroke = 1.5
    ) +
    # étiquettes "Index" et "Observed"
    annotate("text", x = nodes$x[1],
             y = nodes$y[1] + 0.2, label = "Index",
             size = 2.8, color = color, fontface = "bold") +
    annotate("text", x = tail(nodes$x, 1),
             y = tail(nodes$y, 1) + 0.2, label = "Observed",
             size = 2.8, color = color, fontface = "bold") +
    # ---- densité ----
    geom_area(
      data = dens_panel,
      aes(x = x, y = y *2 + 0.0),   # mise à l'échelle pour le panneau
      fill = color, alpha = 0.25
    ) +
    geom_line(
      data = dens_panel,
      aes(x = x, y = y * 2 + 0.0),
      color = color, linewidth = 1
    ) +
    # ligne verticale sur la moyenne (sauf CP)
    {if (!is.na(mean_val))
      geom_vline(xintercept = mean_val,
                 linetype = "dashed", color = color, linewidth = 0.6)} +
    labs(title = route_label, x = "ICC (days)", y = NULL) +
    scale_x_continuous(limits = c(-3, 38)) +
    scale_y_continuous(limits = c(-0.05, 0.95)) +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.y        = element_blank(),
      axis.ticks.y       = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.title         = element_text(face = "bold", color = color, size = 10)
    )
  p
}

p_CP <- make_route_panel("Co-Primary (CP)  —  ICC ≈ 0",
                         n_asymp = -1, color = COL_COPRIM,
                         mu_val = mu_ex, sigma_val = sigma_ex, x_seq = x_seq)
p_PS <- make_route_panel("Primary-Secondary (PS)  —  ICC ≈ SI",
                         n_asymp = 0, color = COL_PRIMARY,
                         mu_val = mu_ex, sigma_val = sigma_ex, x_seq = x_seq)
p_PT <- make_route_panel("Primary-Tertiary (PT)  —  ICC ≈ 2 × SI",
                         n_asymp = 1, color = COL_SECONDARY,
                         mu_val = mu_ex, sigma_val = sigma_ex, x_seq = x_seq)
p_PQ <- make_route_panel("Primary-Quaternary (PQ)  —  ICC ≈ 3 × SI",
                         n_asymp = 2, color = COL_QUAT,
                         mu_val = mu_ex, sigma_val = sigma_ex, x_seq = x_seq)

fig3 <- (p_CP | p_PS) / (p_PT | p_PQ) +
  plot_annotation(
    title   = "Figure 3 — The four transmission routes and their expected ICC distributions",
    caption = "Filled nodes: symptomatic cases  |  Open nodes: asymptomatic cases\nDashed line: expected mean ICC for each route",
    theme   = theme(
      plot.title   = element_text(face = "bold", size = 13),
      plot.caption = element_text(color = "grey50", size = 8)
    )
  )

fig3
```

The observed ICC density is therefore a **mixture** of these components:

$$f_{\text{ICC}}(x) = w_{\text{CP}} \cdot f_{\text{CP}}(x) +
                       w_{\text{PS}} \cdot f_{\text{PS}}(x) +
                       w_{\text{PT}} \cdot f_{\text{PT}}(x) +
                       w_{\text{PQ}} \cdot f_{\text{PQ}}(x)$$

where the $w_k$ are the route weights (with $\sum w_k = 1$), and $\mu$, $\sigma$ are the underlying SI parameters to be estimated.

## The EM algorithm

The parameters $(\mu, \sigma, w_{\text{CP}}, w_{\text{PS}}, w_{\text{PT}}, w_{\text{PQ}})$ are estimated using an **Expectation-Maximisation (EM)** algorithm:

1.  **E-step.** For each observation $x_i$, compute the posterior probability that it belongs to each transmission route.
2.  **M-step.** Update $\mu$ and $\sigma$ by maximising the weighted likelihood, and recompute the weights $w_k$.

These two steps are repeated until convergence. For further technical details, refer to the *Quick Start Guide* vignette of the `mitey` package and to @vink2014.

```{r basic-example, eval=FALSE}
# Basic estimation with 4 routes (default behaviour)
result <- si_estim(icc_data, dist = "normal", n_routes = 4)
result$mean   # Estimated mean serial interval (days)
result$sd     # Estimated standard deviation (days)
```


# Accounting for symptom onset date uncertainty

## The problem

The original Vink et al. method assumes that the reported symptom onset date is exact.
Under this assumption, the likelihood of an observed ICC value $d$ is computed by
integrating the component densities over the unit interval $[d, d+1]$ (or
$[d-1, d+1]$ when $d > 0$), reflecting only the rounding to the nearest integer day.

In practice, however, the reported date may be off by several days. If a surveillance
system collects data once a week, any ICC value computed from these dates carries an
uncertainty of up to $\pm \texttt{wind}$ days around the true value.

```{r wind_figure, echo = FALSE, fig.cap = "Figure X — Illustration of the reporting window"}
library(ggplot2)

wind_ex <- 7

df_wind <- data.frame(
  label = c("True onset\n(index)", "True onset\n(secondary)",
             "Reported onset\n(index)", "Reported onset\n(secondary)"),
  x     = c(2, 15, 7, 21),
  y     = c(2, 2, 1, 1),
  type  = c("true", "true", "reported", "reported")
)

ggplot(df_wind) +
  # Vraies dates
  geom_point(data = df_wind[df_wind$type == "true", ],
             aes(x = x, y = y), color = "#2C6FAC", size = 5) +
  # Dates reportées
  geom_point(data = df_wind[df_wind$type == "reported", ],
             aes(x = x, y = y), color = "#E05A2B", size = 5, shape = 17) +
  # Ligne vraie ICC
  geom_segment(aes(x = 2, xend = 15, y = 2, yend = 2),
               arrow = arrow(ends = "both", length = unit(0.2, "cm"),
                             type = "closed"),
               color = "#2C6FAC", linewidth = 1) +
  annotate("text", x = 5, y = 2.2, label = "True ICC = 13 days",
           color = "#2C6FAC", fontface = "bold", size = 3.5) +
  # Ligne ICC reportée
  geom_segment(aes(x = 7, xend = 21, y = 1, yend = 1),
               arrow = arrow(ends = "both", length = unit(0.2, "cm"),
                             type = "closed"),
               color = "#E05A2B", linewidth = 1) +
  annotate("text", x = 8.5, y = 0.8, label = "Reported ICC = 14 days",
           color = "#E05A2B", fontface = "bold", size = 3.5) +
  # Incertitudes
  geom_errorbarh(aes(xmin = 0, xmax = 0 + wind_ex, y = 1.5,
                     height = 0.15), color = "grey50", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = 14, xmax = 14 + wind_ex, y = 1.5,
                     height = 0.15), color = "grey50", linewidth = 0.8) +
  annotate("text", x = wind_ex / 2, y = 1.65,
           label = paste0("wind = ", wind_ex), color = "grey40", size = 3) +
  annotate("text", x = 10 + wind_ex / 2, y = 1.65,
           label = paste0("wind = ", wind_ex), color = "grey40", size = 3) +
  scale_x_continuous(limits = c(-1, 22)) +
  scale_y_continuous(limits = c(0.5, 2.5),
                     breaks = c(1, 2),
                     labels = c("Reported dates", "True dates")) +
  labs(
    title    = "Figure 4 — Reporting delay distorts the observed ICC",
    subtitle = paste0("With wind = ", wind_ex,
                      ", each date can be shifted by up to ", wind_ex,
                      " days"),
    x = "Time (days)", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor   = element_blank(),
    plot.title         = element_text(face = "bold", size = 13),
    plot.subtitle      = element_text(color = "grey40", size = 10)
  )
```

## The censoring model

To account for this uncertainty, the `mitey` package extends the integration bounds
of the component likelihood functions. Instead of integrating over a window of width 1
(one day, reflecting integer rounding only), the functions now integrate over a window
of width `wind`:

$$\ell_k(d) = \int_{d - \texttt{wind}}^{d} f_k^{\text{lower}}(x, d)\, dx +
              \int_{d}^{d + \texttt{wind}} f_k^{\text{upper}}(x, d)\, dx
\quad \text{(when } d > 0\text{)}$$

$$\ell_k(0) = \int_{0}^{\texttt{wind}} f_k^0(x)\, dx
\quad \text{(when } d = 0\text{)}$$

where $f_k^{\text{lower}}$, $f_k^{\text{upper}}$, and $f_k^0$ are the component
density functions from `flower()`, `fupper()`, and `f0()` respectively, and $k$
indexes the transmission route.

The default value `wind = 1` recovers the original Vink et al. behaviour exactly.

## Usage

The `wind` parameter is available in `si_estim()`, and is automatically forwarded to
all internal integration functions.

```{r wind-usage, eval = FALSE}
# Default behaviour (wind = 1): exact date assumed
result_default <- si_estim(
  dat  = icc_data,
  dist = "normal",
  wind = 1          # default — equivalent to original Vink method
)

# Weekly reporting cycle (wind = 7)
result_wind7 <- si_estim(
  dat  = icc_data,
  dist = "normal",
  wind = 7
)

# No weekend reporting (wind = 2)
result_wind2 <- si_estim(
  dat  = icc_data,
  dist = "normal",
  wind = 2
)
```

As a rule of thumb:

| Surveillance system | Recommended `wind` |
|---|---|
| Daily reporting, exact date | 1 (default) |
| No weekend reporting | 2–3 |
| Twice-weekly reporting | 3–4 |
| Weekly reporting | 7 |
| Bi-weekly reporting | 14 |


------------------------------------------------------------------------

# The key limitation: an arbitrarily fixed number of routes

## The problem

In the original formulation of Vink et al., the number of transmission routes is **fixed at four** (CP, PS, PT, PQ). This choice is reasonable for many common infectious diseases, where it is unlikely to observe more than two or three asymptomatic intermediates between the index case and the observed case.

However, this choice is **arbitrary** and can be ill-suited in two contrasting situations.

## Highly asymptomatic diseases: the risk of underestimating the number of routes

For some diseases, a substantial fraction of infections remains asymptomatic. A notable example is **scabies** (*Sarcoptes scabiei*), for which the incubation period can last several weeks, and infested individuals can be contagious without showing any visible symptoms for a long time.

In this context, long transmission chains are plausible: the index case may have infected one asymptomatic intermediate, who infected another, who infected yet another, before the observed case finally shows symptoms.

```{r chain_with_6_routes , echo = FALSE}
# Chaîne avec 4 intermédiaires asymptomatiques (route 6)
n_asymp_long <- 4
nodes_long   <- make_chain(n_asymp = n_asymp_long, y_base = 1,
                           x_start = 0, x_step = 2.2)
arrows_long  <- make_arrows(nodes_long)

# Accolade ICC totale
icc_total_x  <- c(nodes_long$x[1], tail(nodes_long$x, 1))

fig4 <- ggplot() +
  # Flèches de transmission
  geom_segment(
    data = arrows_long,
    aes(x = x, xend = xend, y = y, yend = yend),
    arrow = arrow(length = unit(0.2, "cm"), type = "closed"),
    color = "grey50", linewidth = 0.9
  ) +
  # Nœuds symptomatiques
  geom_point(
    data = nodes_long %>% filter(symp),
    aes(x = x, y = y),
    color = COL_PRIMARY, size = 8
  ) +
  # Nœuds asymptomatiques
  geom_point(
    data = nodes_long %>% filter(!symp),
    aes(x = x, y = y),
    color = COL_ASYMP, size = 8, shape = 1, stroke = 2
  ) +
  # Étiquettes nœuds
  geom_text(
    data = nodes_long,
    aes(x = x, y = y+0.1,
        label = ifelse(symp,
                       ifelse(x == min(x), "Index", "Observed"),
                       "?")),
    color = ifelse(nodes_long$symp, "white", "grey40"),
    fontface = "bold", size = 3.5
  ) +
  # Légendes sous chaque nœud
  geom_text(
    data = nodes_long,
    aes(x = x, y = y - 0.22,
        label = ifelse(symp,
                       ifelse(x == min(x), "Symptomatic\n(index)", "Symptomatic\n(observed)"),
                       "Asymptomatic\n(unobserved)")),
    size = 2.8,
    color = ifelse(nodes_long$symp, COL_PRIMARY, "grey55"),
    lineheight = 0.85
  ) +
  # Double flèche ICC
  geom_segment(
    aes(x = icc_total_x[1], xend = icc_total_x[2],
        y = 1.45, yend = 1.45),
    arrow = arrow(ends = "both", length = unit(0.2, "cm"), type = "closed"),
    color = COL_ROUTE5, linewidth = 1.3
  ) +
  annotate(
    "text",
    x = mean(icc_total_x), y = 1.58,
    label = paste0("Observed ICC  ≈  5 × SI   (", n_asymp_long,
                   " asymptomatic intermediates)"),
    color = COL_ROUTE5, fontface = "bold", size = 3.8
  ) +
  # Accolades SI individuels (petites flèches sous la chaîne)
  {
    si_arrows <- data.frame(
      x    = head(nodes_long$x, -1),
      xend = tail(nodes_long$x, -1)
    )
    geom_segment(
      data = si_arrows,
      aes(x = x + 0.1, xend = xend - 0.1, y = 0.65, yend = 0.65),
      arrow = arrow(ends = "both", length = unit(0.12, "cm"), type = "open"),
      color = "grey60", linewidth = 0.6
    )
  } +
  annotate(
    "text", x = mean(nodes_long$x), y = 0.52,
    label = "Each arrow ≈ 1 serial interval (SI)",
    color = "grey55", size = 3, fontface = "italic"
  ) +
  scale_x_continuous(limits = c(-1.2, max(nodes_long$x) + 1.5)) +
  scale_y_continuous(limits = c(0.3, 1.75)) +
  labs(
    title   = "Figure 5 — Long transmission chain in a highly asymptomatic disease",
    subtitle = "The observed ICC spans 5 serial intervals, but only 2 symptomatic cases are recorded",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid       = element_blank(),
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(color = "grey40", size = 10)
  )

fig4


```

If such long chains are frequent but the algorithm is limited to 4 routes, the PT or PQ component will absorb ICC values that actually correspond to routes of order 5 or 6. This leads to **overestimation of the SI**, because the algorithm assigns an ICC of $5\mu$ to the component centred at $3\mu$.

## Highly symptomatic diseases: the risk of over-parameterisation

Conversely, for a disease where almost all infections are symptomatic (e.g. measles or smallpox), it is unlikely that several asymptomatic intermediates follow one another in a single transmission chain. In this case, **modelling 4 routes introduces unnecessary parameters** (the weights for PT and PQ routes will be estimated near zero), at the cost of precision and numerical stability of the EM algorithm.

A model with only 2 or 3 routes would be more parsimonious and would yield more stable estimates.

## Summary

```{r routes-table, echo=FALSE}
df_routes <- data.frame(
  Context = c(
    "Highly asymptomatic disease",
    "Moderately asymptomatic disease",
    "Highly symptomatic disease"
  ),
  Examples = c(
    "Hepatitis A",
    "Influenza",
    "Measles, smallpox"
  ),
  `Recommended routes` = c(
    "\u2265 5 \u2013 8",
    "3 \u2013 5 (default = 4)",
    "2 \u2013 3"
  ),
  `Risk if n_routes too low` = c(
    "Overestimation of the SI",
    "Low",
    "\u2013"
  ),
  `Risk if n_routes too high` = c(
    "\u2013",
    "Low",
    "Over-parameterisation, instability"
  )
)
knitr::kable(
  df_routes,
  caption = "Impact of the number of routes depending on the epidemiological profile of the disease",
  col.names = c("Context", "Examples", "Recommended routes",
                "Risk if too low", "Risk if too high")
)
```

------------------------------------------------------------------------

# The new feature: `n_routes` as a user parameter

## Principle

The central modification brought to `mitey` is to **expose the number of transmission routes as an explicit parameter** of `si_estim()`, rather than hard-coding it to 4. The user can now specify any integer $\geq 2$.

The default value remains `n_routes = 4` to ensure backward compatibility with the original behaviour.

## What changed in the code

The change affects several functions in a consistent way:

| Function | Role | Modification |
|---|---|---|
| `si_estim()` | Main estimation function | `n_routes` and `wind` parameters added |
| `f_norm()` | Mixture density (normal distribution) | Adapted for $n$ components |
| `f_gam()` | Mixture density (gamma distribution) | Adapted for $n$ components |
| `f0()` | Zero-ICC component density | Integration bound extended to `wind` |
| `flower()` | Lower-bound component density | Integration bound extended to `wind` |
| `fupper()` | Upper-bound component density | Integration bound extended to `wind` |
| `integrate_component()` | Numerical integration per component | Bounds adapted to `wind` |
| `integrate_components_wrapper()` | Integration across all components | `n_routes` and `wind` forwarded |
| `plot_si_fit()` | Fitted distribution visualisation | Adapted for $n$ components |
| `plot_si_fit_result()` | Visualisation wrapper | `n_routes` forwarded |
| `compare_n_routes()` | Model comparison across route counts | `wind` forwarded to `si_estim()` |

The component numbering logic is as follows (normal distribution):

-   Component 1: Co-Primary route (half-normal)
-   Components $2i$ and $2i+1$ for $i = 1, \ldots, n_{\text{routes}} - 1$: positive and negative routes for transmission route $i+1$

For the gamma distribution (which only produces positive values), only even components are used, giving $n_{\text{routes}}$ components in total.

## Updated function signatures

```{r signatures, eval=FALSE}
# Estimation with a custom number of routes
result <- si_estim(
  dat      = icc_data,
  dist     = "normal",   # or "gamma"
  n_routes = 6L          # desired number of routes (default: 4)
)

# Direct visualisation of the result
plot_si_fit_result(result, dat = icc_data, dist = "normal")
```

The function returns the same fields as before, with two additions:

-   `n_routes`: the number of routes used (useful for traceability)
-   `loglik`: the log-likelihood of the fitted model (useful for model comparison, see §7)

------------------------------------------------------------------------

# Choosing the right number of routes: `compare_n_routes()`

## Model comparison principle

Increasing the number of routes is more likely to improve the fit to the data (the log-likelihood can only increase or stay the same), but at the cost of **greater complexity**. A **penalised criterion** is therefore needed to find the right balance.

The package implements two classical criteria:

$$\text{AIC} = -2\ell + 2p \qquad \text{BIC} = -2\ell + p \ln(n)$$

where $\ell$ is the log-likelihood, $p$ the number of free parameters, and $n$ the sample size. The BIC penalises complexity more heavily and therefore favours more parsimonious models --- it is generally recommended for selecting the number of components in a mixture model.

The number of free parameters is: - **Normal distribution:** $p = 2 + (2 \times n_{\text{routes}} - 2)$ (2 for $\mu$ and $\sigma$, plus the free weights) - **Gamma distribution:** $p = 2 + (n_{\text{routes}} - 1)$

## The `compare_n_routes()` function

```{r compare-n-routes-example, eval=FALSE}
# Compare models from 2 to 8 routes
comparison <- compare_n_routes(
  dat          = icc_data,
  n_routes_max = 8L,
  dist         = "normal",
  n            = 50        # number of EM iterations per model
)
```

The function returns a summary table and automatically displays the fitted distribution plots for each value of `n_routes` tested:

```{r compare-n-routes-output, eval=FALSE, echo = TRUE}
# Structure of the output
# n_routes | mu   | sigma | loglik | aic    | bic    | n_params | converged | iterations
# 2        | ...  | ...   | ...    | ...    | ...    | ...      | TRUE      | ...
# 3        | ...  | ...   | ...    | ...    | ...    | ...      | TRUE      | ...
# ...
```

## Starting heuristics: `n_optimal1()`, `n_optimal2()`

Two simple data-driven heuristics are available to guide the initial choice of `n_routes_max`:

```{r heuristics, eval=FALSE}
# Heuristic 1: based on the 95th percentile and the mean
n_optimal1(icc_data)   # as.integer(quantile(dat, 0.95) / mean(dat))

# Heuristic 2: based on the relative range
n_optimal2(icc_data)   # as.integer(max(dat) / sd(dat))
```

These heuristics are not rigorous estimators but reasonable **starting points** for exploring the `n_routes` grid. It is recommended to use them to set `n_routes_max` in `compare_n_routes()`, and then rely on AIC/BIC for the final selection.

------------------------------------------------------------------------

# Applied examples

## Example 1 --- Highly asymptomatic disease (simulated data)

We simulate a disease with a high probability of asymptomatic infection. This results in ICC values that may correspond to long transmission chains: some observed ICCs are 5× or 6× the true SI.

```{r simulation-high-asymptomatic, fig5, echo = FALSE}
set.seed(42)

true_mu_asymp    <- 10   # Vrai IS moyen : 10 jours
true_sigma_asymp <- 2    # Vrai écart-type : 2 jours
N          <- 300  # Nombre total d'observations

# Poids des 6 routes
# (CP, PS, PT, PQ, Route-5, Route-6)

route_weights <- c(
  CP = 0.10,
  PS = 0.34,
  PT = 0.23,
  PQ = 0.13,
  R5 = 0.12,
  R6 = 0.08
)
# Nombre de tirages par route
n_per_route <- round(route_weights * N)

# Route 1 — Co-Primary : demi-normale
CP <- fdrtool::rhalfnorm(
  n_per_route["CP"],
  theta = sqrt(pi / 2) / (sqrt(2) * true_sigma_asymp)
)

# Route 2 — Primary-Secondary : N(mu, sigma^2)
PS <- rnorm(n_per_route["PS"],
            mean = true_mu_asymp,
            sd   = true_sigma_asymp)

# Route 3 — Primary-Tertiary : N(2*mu, 2*sigma^2)
PT <- rnorm(n_per_route["PT"],
            mean = 2 * true_mu_asymp,
            sd   = sqrt(2) * true_sigma_asymp)

# Route 4 — Primary-Quaternary : N(3*mu, 3*sigma^2)
PQ <- rnorm(n_per_route["PQ"],
            mean = 3 * true_mu_asymp,
            sd   = sqrt(3) * true_sigma_asymp)

# Route 5 : N(4*mu, 4*sigma^2)
R5 <- rnorm(n_per_route["R5"],
            mean = 4 * true_mu_asymp,
            sd   = sqrt(4) * true_sigma_asymp)

# Route 6 : N(5*mu, 5*sigma^2)
R6 <- rnorm(n_per_route["R6"],
            mean = 5 * true_mu_asymp,
            sd   = sqrt(5) * true_sigma_asymp)

# Assemblage et arrondi à l'entier (jours)
icc_asymp <- round(c(CP, PS, PT, PQ, R5, R6))
icc_asymp<-abs(icc_asymp)

# Vérification rapide
cat("=== Simulation summary ===\n")
cat("N observations :", length(icc_asymp), "\n")
cat("True SI mean   :", true_mu_asymp, "days\n")
cat("True SI sd     :", true_sigma_asymp, "days\n")
cat("ICC range      :", min(icc_asymp), "to", max(icc_asymp), "days\n")
cat("ICC mean       :", round(mean(icc_asymp), 2), "days\n\n")
cat("Observations per route:\n")
print(n_per_route)

# Histogramme de vérification
hist_sim <- ggplot(data.frame(icc = icc_asymp), aes(x = icc)) +
  geom_histogram(
    binwidth = 1, fill = "lightsteelblue", color = "white", linewidth = 0.3
  ) +
  geom_vline(xintercept = true_mu_asymp,       linetype = "dashed",
             color = COL_TRUE_SI, linewidth = 1) +
  geom_vline(xintercept = 2 * true_mu_asymp,   linetype = "dashed",
             color = COL_TRUE_SI, linewidth = 0.7, alpha = 0.6) +
  geom_vline(xintercept = 3 * true_mu_asymp,   linetype = "dashed",
             color = COL_TRUE_SI, linewidth = 0.7, alpha = 0.6) +
  geom_vline(xintercept = 4 * true_mu_asymp,   linetype = "dashed",
             color = COL_TRUE_SI, linewidth = 0.7, alpha = 0.4) +
  geom_vline(xintercept = 5 * true_mu_asymp,   linetype = "dashed",
             color = COL_TRUE_SI, linewidth = 0.7, alpha = 0.4) +
  annotate("text", x = true_mu_asymp + 0.5, y = Inf,
           label = "SI", color = COL_TRUE_SI,
           vjust = 2, hjust = 0, size = 3.5, fontface = "italic") +
  labs(
    title    = "Simulated ICC data — 6 transmission routes",
    subtitle = paste0("True SI: mean = ", true_mu_asymp, " days, sd = ", true_sigma_asymp,
                      " days  |  N = ", length(icc_asymp), " observations"),
    x = "ICC interval (days)",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title    = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(color = "grey40", size = 10))

hist_sim
```

The original code with 4 routes will find a density function as below.

```{r example_asymptomatic, echo = FALSE}

result_asymp_4 <-si_estim(icc_asymp)
graph<- plot_si_fit_result(result_asymp_4,icc_asymp)+
  geom_vline(
    xintercept = true_mu_asymp,
    color = COL_TRUE_SI, linetype = "dotted", linewidth = 1
  ) +
  annotate("text",
           x = true_mu_asymp - 7, y = Inf,
           label = paste0("True SI = ", true_mu_asymp, " days"),
           color = COL_TRUE_SI, vjust = 2, hjust = 0, size = 3.2) +
  labs(
    title    = "Fit with 4 routes (default)",
    subtitle = paste0("Estimated mean = ", round(result_asymp_4$mean, 2),
                      " days  |  True SI = ", true_mu_asymp, " days")
  )
graph

```
The fourth route looks overestimated, which is because all the ICC intervals whose values are greater than $4\mu$ are assigned by the algorithm to the fourth route. This increases the fourth route weight.

The compare_n_routes used below shows how setting the number of routes helps the algorithm capture the distribution.

```{r compare-routes-asym, eval=TRUE, echo = FALSE}
# Formal comparison by AIC/BIC
comparison_asym <- compare_n_routes(
  dat          = icc_asymp,
  n_routes_max = 7L,
  dist         = "normal"
)
```

These graphs show how increasing the number of routes has splitted all ICC intervals in the right component they belong to.


## Example 2 --- Highly symptomatic disease (simulated data)

We now simulate a disease where almost all cases are symptomatic: long routes (PT, PQ) are very rare.

```{r symptomatic-simulation, echo = FALSE, eval = TRUE}
# --- Simulation : maladie très symptomatique (2 routes suffisent) ---
set.seed(123)

true_mu_symp    <- 20    # Vrai IS : 20 jours
true_sigma_symp <- 3 # Vrai écart-type : 17 jours
N_symp          <- 300

# Poids : quasi-exclusivement CP et PS, PT anecdotique
n_symp <- round(c(
  CP = 0.25,
  PS = 0.75,
  PT = 0.00
) * N_symp)

icc_symp <- round(c(
  fdrtool::rhalfnorm(
    n_symp["CP"],
    theta = sqrt(pi / 2) / (sqrt(2) * true_sigma_symp)
  ),
  rnorm(n_symp["PS"],
        mean = true_mu_symp,
        sd   = true_sigma_symp),
  rnorm(n_symp["PT"],
        mean = 2 * true_mu_symp,
        sd   = sqrt(2) * true_sigma_symp)
))

cat("True SI mean :", true_mu_symp, "days\n")
cat("N observations :", length(icc_symp), "\n")
cat("ICC range :", min(icc_symp), "to", max(icc_symp), "days\n")
```

```{r symptomatic-fit-4routes, echo = FALSE, eval = FALSE}
# --- Estimation avec 4 routes (comportement par défaut) ---
result_symp_4 <- si_estim(
  dat      = icc_symp,
  dist     = "normal",
  n_routes = 4L,
  n        = 50
)

cat("=== 4 routes ===\n")
cat("Estimated mean :", round(result_symp_4$mean, 2), "days\n")
cat("Estimated sd   :", round(result_symp_4$sd,   2), "days\n")
cat("Converged      :", result_symp_4$converged, "\n")
cat("Component weights:\n")
print(round(result_symp_4$wts, 3))

plot_4 <- plot_si_fit_result(
  si_result = result_symp_4,
  dat       = icc_symp,
  dist      = "normal"
) +
  geom_vline(
    xintercept = true_mu_symp,
    color = COL_TRUE_SI, linetype = "dotted", linewidth = 1
  ) +
  annotate("text",
           x = true_mu_symp + 0.3, y = Inf,
           label = paste0("True SI = ", true_mu_symp, " days"),
           color = COL_TRUE_SI, vjust = 2, hjust = 0, size = 3.2) +
  labs(
    title    = "Fit with 4 routes (default)",
    subtitle = paste0("Estimated mean = ", round(result_symp_4$mean, 2),
                      " days  |  True SI = ", true_mu_symp, " days")
  )
plot_4
```

```{r symptomatic-fit-3 routes, echo = FALSE, eval = FALSE}
# --- Estimation avec 3 routes (modèle parcimonieux) ---
result_symp_3 <- si_estim(
  dat      = icc_symp,
  dist     = "normal",
  n_routes = 3L,
  n        = 50
)

cat("=== 3 routes ===\n")
cat("Estimated mean :", round(result_symp_3$mean, 3), "days\n")
cat("Estimated sd   :", round(result_symp_3$sd,   3), "days\n")
cat("Converged      :", result_symp_3$converged, "\n")
cat("Component weights:\n")
print(round(result_symp_3$wts, 3))

plot_3 <- plot_si_fit_result(
  si_result = result_symp_3,
  dat       = icc_symp,
  dist      = "normal"
) +
  geom_vline(
    xintercept = true_mu_symp,
    color = COL_TRUE_SI, linetype = "dotted", linewidth = 1
  ) +
  annotate("text",
           x = true_mu_symp + 0.3, y = Inf,
           label = paste0("True SI = ", true_mu_symp, " days"),
           color = COL_TRUE_SI, vjust = 2, hjust = 0, size = 3.2) +
  labs(
    title    = "Fit with 3 routes (parsimonious)",
    subtitle = paste0("Estimated mean = ", round(result_symp_3$mean, 2),
                      " days  |  True SI = ", true_mu_symp, " days")
  )

plot_3
```

```{r compare-routes-sym, eval=TRUE, echo = FALSE}
# Formal comparison by AIC/BIC
comparison_sym <- compare_n_routes(
  dat          = icc_symp,
  n_routes_max = 4L,
  dist         = "normal"
)
```

The example above shows us that the initial algorithm is robust to the case where the number of routes set in the algorithm is greater that the number of actual routes. Then, the algorihm will find weights close to 0.

## Example 3 --- Example with real outbreak data - Influenza in Canada

We now run the simulation using real outbreak data and compare the results from 4 to 8 routes.

```{r Influenza Canada real data recreation, echo = FALSE}
icc_Influenza_Canada<-c(0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 3, 4, 4, 4, 4, 4, 4, 5, 5, 5, 5, 6, 6, 6, 7, 8, 8, 9, 14, 16, 17, 20)
```

The example below uses the data from outbreak of Influenza A(H1N1)pdm09 in the study from Savage, in 2011 in Canada.

```{r Influenza Canada 4 routes, echo = FALSE}
# --- Estimation avec 4 routes (comportement par défaut) ---
result_Influenza <- si_estim(
  dat      = icc_Influenza_Canada,
  dist     = "normal",
  n_routes = 4L,
  n        = 50
)

cat("=== 4 routes ===\n")
cat("Estimated mean :", round(result_Influenza$mean, 2), "days\n")
cat("Estimated sd   :", round(result_Influenza$sd,   2), "days\n")
cat("Converged      :", result_Influenza$converged, "\n")
cat("Component weights:\n")
print(round(result_Influenza$wts, 3))

plot_influenza <- plot_si_fit_result(
  si_result = result_Influenza,
  dat       = icc_Influenza_Canada,
  dist      = "normal"
) +
  labs(
    title    = "Fit with 4 routes (default)",
    subtitle = paste0("Estimated mean = ", round(result_Influenza$mean, 2)," days")
  )
plot_influenza
```
In the original version of the algorithm, the ICC intervals whose values are greater than $4\mu$ are captured by the fourth route.

With the new version of the algorithm, those last value are well captured by the 8 routes : the weight for the fourth route has decreased, and the ICC intervals are splitted in the route they belong to. 

```{r Influenza Canada 8 routes, echo = FALSE}
# --- Estimation avec 8 routes ---
result_Influenza_8 <- si_estim(
  dat      = icc_Influenza_Canada,
  dist     = "normal",
  n_routes = 8L,
  n        = 50
)

cat("=== 8 routes ===\n")
cat("Estimated mean :", round(result_Influenza_8$mean, 2), "days\n")
cat("Estimated sd   :", round(result_Influenza_8$sd,   2), "days\n")
cat("Converged      :", result_Influenza_8$converged, "\n")
cat("Component weights:\n")
print(round(result_Influenza_8$wts, 3))

plot_influenza_8 <- plot_si_fit_result(
  si_result = result_Influenza_8,
  dat       = icc_Influenza_Canada,
  dist      = "normal"
) +
  labs(
    title    = "Fit with 8 routes",
    subtitle = paste0("Estimated mean = ", round(result_Influenza$mean, 2)," days")
  )
plot_influenza_8
```

## Example 4 --- Example with real outbreak data - Pertussis in Netherlands

```{r Pertussis Netherlands Greef, echo = FALSE}
icc_pertussis_netherlands<-c(0,0,0,0,1,2,2,2,2,3,3,3,3,3,4,4,4,5,5,5,5,5,5,5,6,6,6,6,6,6,7,7,7,7,7,7,8,8,8,8,8,9,9,9,9,9,10,10,10,10,10,10,10,10,10,10,11,11,11,11,12,12,12,12,12,12,12,12,12,13,13,13,13,13,13,13,13,13,14,14,14,14,14,14,15,15,15,15,15,15,15,15,15,16,16,16,16,16,16,16,16,17,17,17,17,17,17,18,18,18,18,18,18,18,18,19,19,19,19,19,19,19,19,19,20,20,20,20,20,20,20,20,20,20,20,21,21,21,21,21,21,22,22,22,22,22,22,22,22,23,23,23,23,23,23,23,23,24,24,24,24,24,24,24,24,25,25,25,25,25,25,25,26,26,26,27,27,27,27,27,28,28,28,28,28,28,28,28,28,28,28,29,29,29,29,29,29,30,30,30,30,30,30,30,30,30,30,31,31,31,31,31,31,32,32,32,32,33,33,33,33,33,33,33,34,34,34,35,35,35,35,36,36,36,36,37,37,37,37,37,38,38,40,40,40,40,41,41,41,41,42,42,42,42,42,42,42,43,44,45,46,47,47,47,47,47,48,48,49,49,50,51,51,51,52,52,52,53,53,53,54,54,55,56,56,57,58,59,60,61,62,64,64,65,66,67,67,68,70,72,76,77,80,81,84,90,92,93,101,105,108,116,116,130,155)
```

Using the dataset from a pertussis outbreak in Netherlands, from Greeff's study in 2010, the last values are not captured by the first algorithm.

```{r Pertussis Netherlands 4 routes, echo = FALSE}
# --- Estimation avec 4 routes (comportement par défaut) ---
result_pertussis_4 <- si_estim(
  dat      = icc_pertussis_netherlands,
  dist     = "normal",
  n_routes = 4L,
  n        = 50
)

cat("=== 4 routes ===\n")
cat("Estimated mean :", round(result_pertussis_4$mean, 2), "days\n")
cat("Estimated sd   :", round(result_pertussis_4$sd,   2), "days\n")
cat("Converged      :", result_pertussis_4$converged, "\n")
cat("Component weights:\n")
print(round(result_pertussis_4$wts, 3))

plot_pertussis <- plot_si_fit_result(
  si_result = result_pertussis_4,
  dat       = icc_pertussis_netherlands,
  dist      = "normal"
) +
  labs(
    title    = "Fit with 4 routes (default)",
    subtitle = paste0("Estimated mean = ", round(result_pertussis_4$mean, 2)," days")
  )
plot_pertussis
```

With the new version of the algorithm, those last value are captured better by the 8 routes.

```{r Pertussis Netherlands 8 routes, echo = FALSE}
# --- Estimation avec 8 routes ---
result_Pertussis_8 <- si_estim(
  dat      = icc_pertussis_netherlands,
  dist     = "normal",
  n_routes = 8L,
  n        = 50
)

cat("=== 8 routes ===\n")
cat("Estimated mean :", round(result_Pertussis_8$mean, 2), "days\n")
cat("Estimated sd   :", round(result_Pertussis_8$sd,   2), "days\n")
cat("Converged      :", result_Pertussis_8$converged, "\n")
cat("Component weights:\n")
print(round(result_Pertussis_8$wts, 3))

plot_pertussis_8 <- plot_si_fit_result(
  si_result = result_Pertussis_8,
  dat       = icc_pertussis_netherlands,
  dist      = "normal"
) +
  labs(
    title    = "Fit with 8 routes",
    subtitle = paste0("Estimated mean = ", round(result_Pertussis_8$mean, 2)," days")
  )
plot_pertussis_8
```



------------------------------------------------------------------------

# Conclusion and practical recommendations

## Summary

This vignette has presented the extension brought to the `si_estim()` function of the `mitey` package, which now allows the user to freely specify the number of transmission routes $n_{\text{routes}}$ and an uncertainty window of the recorded dates.

The key takeaways are:

-   The number of routes is a **modelling choice** that should be guided by the biology of the disease.
-   Using **too few routes** for a highly asymptomatic disease leads to overestimation of the SI.
-   Using **too many routes** for a highly symptomatic disease introduces unnecessary parameters and can destabilise the EM algorithm.
-   The `compare_n_routes()` function automates the comparison and allows the optimal model to be selected objectively.

## Recommended workflow

```{r recommended-workflow, eval=FALSE}
# Step 1: get a ballpark for the number of routes using the heuristics
n1 <- n_optimal1(icc_data)
n2 <- n_optimal2(icc_data)


cat("Heuristics:", n1, n2, "\n")
n_max_explore <- max(n1, n2) + 2   # safety margin

# Step 2: compare models from 2 to n_max_explore routes
comparison <- compare_n_routes(
  dat          = icc_data,
  n_routes_max = n_max_explore,
  dist         = "normal"
)

# Step 3: estimate with the BIC-optimal number of routes
n_opt <- comparison$n_routes[which.min(comparison$bic)]
result_final <- si_estim(icc_data, dist = "normal", n_routes = n_opt)

# Step 4: visualise the final fit
plot_si_fit_result(result_final, dat = icc_data, dist = "normal")
```


------------------------------------------------------------------------

# References

::: {#refs}
:::

------------------------------------------------------------------------

*This document was generated with `rmarkdownr packageVersion("rmarkdown")` and `miteyr packageVersion("mitey")`.*
