---
title: "Accounting for Date Uncertainty with the `wind` Parameter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Accounting for Date Uncertainty with the `wind` Parameter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE
)
library(mitey)
library(extraDistr)
```

## 1. The `mitey` package in brief

The `mitey` package implements the Vink et al. (2014) method for estimating the
**serial interval** of an infectious disease from outbreak data. The serial
interval is defined as the time between symptom onset in a primary (index) case
and symptom onset in a secondary case infected by that primary case.

In practice, what is recorded during an outbreak is not the serial interval
itself but the **index-case-to-case (ICC) interval**: the observed time between
the symptom onset date of the index case and the symptom onset date of the
secondary case. The core function `si_estim()` fits a mixture model to a vector
of ICC intervals and returns estimates of the mean $\mu$ and standard deviation
$\sigma$ of the underlying serial interval distribution, along with the weights
of the different transmission routes (co-primary, primary-secondary,
primary-tertiary, etc.).

---

## 2. The problem: date uncertainty propagates to interval uncertainty

In many outbreak investigations, symptom onset dates are not recorded with
daily precision. A surveillance system may aggregate reports **weekly** or surveys might be done during the weekdays. In such settings,
what is known is not the exact day of symptom onset, but only that it occurred
**within a given observation window**.

## 2.1. A worked example: weekday-only recording

As a concrete illustration, suppose the index case has a symptom onset
drawn uniformly between day 1 and day 14, and the secondary case develops
symptoms a number of days later drawn from a normal distribution with
mean $\mu = 22$ and standard deviation $\sigma = 3$. Under daily
recording, the ICC interval $I = P_2 - P_1$ exactly follows this normal
distribution. Under weekday-only recording, onset dates falling on
weekends are pushed forward to the next Monday, distorting both the
marginal distributions of $P_1$ and $P_2$ and, crucially, the
distribution of their difference.

```{r}
library(extraDistr)

N     <- 10000
mu    <- 22
sigma <- 3

# P1: index case onset, uniformly distributed between day 1 and 14
P1   <- rdunif(n = N, min = 1, max = 14)

# norm: true serial interval drawn from a normal distribution
norm <- trunc(rnorm(n = N, mean = mu, sd = sigma))

# P2: secondary case onset
P2 <- P1 + norm

# I: true ICC interval
I <- P2 - P1

breaks1 <- seq(0.5, 15.5, 1)

par(mfrow = c(1, 3))
hist(P1, main = "P1: index case onset\n(Uniform)", xlab = "Days",
     col = "skyblue", breaks = breaks1)
hist(P2, main = "P2: secondary case onset\n(P1 + SI)", xlab = "Days",
     col = "lightgreen")
hist(I,  main = "True ICC interval\n(I = P2 - P1)", xlab = "Days",
     col = "cyan")
par(mfrow = c(1, 1))
```



Now suppose that symptom onset dates are only recorded on weekdays.
Dates falling on a Saturday (day $\equiv 6 \pmod 7$) or Sunday
(day $\equiv 0 \pmod 7$) are instead recorded on the following Monday.
This shifts $P_1$ and $P_2$ forward by up to two days, and the
resulting observed ICC interval $I' = P_2' - P_1'$ can differ
substantially from the true interval $I$.

```{r}
# Shift weekend dates to the following Monday
shift_to_weekday <- function(x) {
  remainder <- x %% 7
  ifelse(remainder == 0 & x > 3, x + 1,
  ifelse(remainder == 6 & x > 3, x + 2, x))
}

P1_prime <- shift_to_weekday(P1)
P2_prime <- shift_to_weekday(P2)
I_prime  <- P2_prime - P1_prime

par(mfrow = c(1, 3))
hist(P1_prime, main = "P1': recorded onset\n(weekdays only)", xlab = "Days",
     col = "skyblue", breaks = breaks1)
hist(P2_prime, main = "P2': recorded onset\n(weekdays only)", xlab = "Days",
     col = "lightgreen")
hist(I_prime,  main = "Observed ICC interval\n(I' = P2' - P1')", xlab = "Days",
     col = "cyan")
par(mfrow = c(1, 1))
```

Comparing the two distributions makes the problem clear: the weekday
recording rule creates gaps and spikes in the observed ICC distribution
that have nothing to do with the underlying biology. Feeding these
shifted intervals directly into `si_estim()` without correction will
bias the estimates of $\mu$ and $\sigma$.


## 3. Solution: the `wind` parameter

### Idea

Instead of treating each observed value $d$ as a point observation, `mitey`
treats it as the **centre of an uncertainty window** of width `wind`. The true
ICC interval is assumed to lie somewhere in $[d - w,\; d + w]$, with
probability decreasing as we move away from $d$.

The likelihood contribution of observation $d$ is therefore not simply
$f_{\text{SI}}(d)$ but the **weighted integral** of the serial interval density
over the window:

$$L(d \mid \mu, \sigma) =
  \int_{d-w}^{d+w} f_{\text{SI}}(x)\; f_\triangle(x \mid d, w)\; dx$$

where $f_\triangle(x \mid d, w)$ is a **triangular weight function** centred at
$d$, equal to 1 at $x = d$ and falling linearly to 0 at $x = d \pm w$.

### Why a triangular weight?

If both $P_1$ and $P_2$ can each fall **uniformly** within a window of width
$w$ around their recorded date, the difference $P_2 - P_1$ follows a
**triangular distribution** on $[-w, w]$ around the true difference. The
triangular weight function directly reflects this uncertainty structure.

### Usage

Simply pass the observation window (in the same unit as your ICC intervals,
typically days) to `si_estim()` via the `wind` argument:

```{r usage, eval = FALSE}
# Without correction (default: wind = 1, point observation)
result_no_corr <- si_estim(icc_data, n = 100)

# With a 7-day observation window
result_corrected <- si_estim(icc_data, n = 100, wind = 7)
```

