---
title: "Simulation Study Execution V1"
author: "Terry Xue"
output:
  pdf_document:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
params:
  n_reps: 100
  n_iter: 50
  n_starts: 1
  seed: 20260706
  progress_every: 10
  scenario_subset: !r NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

current_dir <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
project_root <- if (basename(current_dir) == "meeting_summarys") {
  normalizePath(file.path(current_dir, ".."), winslash = "/", mustWork = TRUE)
} else {
  current_dir
}

local_mitey_path <- file.path(project_root, "mitey", "mitey")
pkgload::load_all(local_mitey_path, quiet = TRUE)
```

# 1. Purpose

This document executes the simulation study described in
`Simulation_Study_Plan_V4.Rmd`. The goal is to evaluate whether the lognormal
extension of the Vink serial interval estimation framework can recover the true
SI mean and SD from simulated ICC intervals.

The fitted model is always:

```{r fitted-method, eval=FALSE}
fit <- si_estim(
  dat = observed_icc_intervals,
  dist = "lognormal",
  n = params$n_iter,
  n_starts = params$n_starts
)
```

The data-generating distribution varies across scenarios. Correct-specification
scenarios generate SI values from a lognormal distribution. Misspecification
scenarios generate SI values from Gamma or Normal distributions but still fit
the lognormal extension.

# 2. Setup

```{r simulation-settings}
n_reps <- params$n_reps
n_iter <- params$n_iter
n_starts <- params$n_starts
seed <- params$seed
progress_every <- params$progress_every

set.seed(seed)

simulation_settings <- data.frame(
  n_reps = n_reps,
  n_iter = n_iter,
  n_starts = n_starts,
  progress_every = progress_every,
  seed = seed
)

simulation_settings
```

# 3. Scenario Table

```{r scenario-table}
scenario_grid <- data.frame(
  scenario_id = c(
    "A0_baseline",
    "A1_n100", "A1_n300", "A1_n600", "A1_n1000",
    "A2_PS_dominant", "A2_CP_heavy", "A2_higher_generation_heavy",
    "B1_gamma", "B2_normal"
  ),
  scenario_type = c(
    rep("correct_specification", 8),
    "misspecification", "misspecification"
  ),
  true_dist = c(
    rep("lognormal", 8),
    "gamma", "normal"
  ),
  n_obs = c(
    300,
    100, 300, 600, 1000,
    300, 300, 300,
    300, 300
  ),
  true_mean = 6,
  true_sd = 3,
  w_cp = c(0.20, 0.20, 0.20, 0.20, 0.20, 0.10, 0.40, 0.10, 0.20, 0.20),
  w_ps = c(0.50, 0.50, 0.50, 0.50, 0.50, 0.70, 0.35, 0.35, 0.50, 0.50),
  w_pt = c(0.20, 0.20, 0.20, 0.20, 0.20, 0.15, 0.15, 0.35, 0.20, 0.20),
  w_pq = c(0.10, 0.10, 0.10, 0.10, 0.10, 0.05, 0.10, 0.20, 0.10, 0.10),
  stringsAsFactors = FALSE
)

if (!is.null(params$scenario_subset)) {
  scenario_grid <- scenario_grid[scenario_grid$scenario_id %in% params$scenario_subset, ]
}

scenario_grid$weights <- sprintf(
  "(%.2f, %.2f, %.2f, %.2f)",
  scenario_grid$w_cp,
  scenario_grid$w_ps,
  scenario_grid$w_pt,
  scenario_grid$w_pq
)

scenario_grid
```

# 4. Simulation Functions

The functions below generate ICC intervals using the Vink path construction.
For lognormal and Gamma SI generation, the path-specific continuous ICC
intervals are

```text
CP = abs(X1 - X2)
PS = X1
PT = X1 + X2
PQ = X1 + X2 + X3
```

For Normal SI generation, the folded Vink-style version is used:

```text
CP = abs(X1 - X2)
PS = abs(X1)
PT = abs(X1 + X2)
PQ = abs(X1 + X2 + X3)
```

Observed ICC intervals are rounded to integer days and truncated at zero.

```{r simulation-functions}
natural_to_lognormal <- function(mean, sd) {
  sdlog <- sqrt(log(1 + sd^2 / mean^2))
  meanlog <- log(mean) - 0.5 * sdlog^2
  c(meanlog = meanlog, sdlog = sdlog)
}

draw_si <- function(n, dist, mean, sd) {
  if (n == 0) {
    return(numeric(0))
  }

  if (dist == "lognormal") {
    pars <- natural_to_lognormal(mean, sd)
    return(rlnorm(n, meanlog = pars[["meanlog"]], sdlog = pars[["sdlog"]]))
  }

  if (dist == "gamma") {
    shape <- mean^2 / sd^2
    scale <- sd^2 / mean
    return(rgamma(n, shape = shape, scale = scale))
  }

  if (dist == "normal") {
    return(rnorm(n, mean = mean, sd = sd))
  }

  stop("Unsupported SI distribution: ", dist)
}

route_counts <- function(n_obs, weights) {
  counts <- setNames(as.integer(round(weights * n_obs)), names(weights))
  counts["PS"] <- n_obs - sum(counts[c("CP", "PT", "PQ")])
  counts
}

simulate_icc_intervals <- function(n_obs, true_dist, true_mean, true_sd, weights) {
  counts <- route_counts(n_obs, weights)

  cp <- abs(
    draw_si(counts[["CP"]], true_dist, true_mean, true_sd) -
      draw_si(counts[["CP"]], true_dist, true_mean, true_sd)
  )

  if (true_dist == "normal") {
    ps <- abs(draw_si(counts[["PS"]], true_dist, true_mean, true_sd))
    pt <- abs(
      draw_si(counts[["PT"]], true_dist, true_mean, true_sd) +
        draw_si(counts[["PT"]], true_dist, true_mean, true_sd)
    )
    pq <- abs(
      draw_si(counts[["PQ"]], true_dist, true_mean, true_sd) +
        draw_si(counts[["PQ"]], true_dist, true_mean, true_sd) +
        draw_si(counts[["PQ"]], true_dist, true_mean, true_sd)
    )
  } else {
    ps <- draw_si(counts[["PS"]], true_dist, true_mean, true_sd)
    pt <- draw_si(counts[["PT"]], true_dist, true_mean, true_sd) +
      draw_si(counts[["PT"]], true_dist, true_mean, true_sd)
    pq <- draw_si(counts[["PQ"]], true_dist, true_mean, true_sd) +
      draw_si(counts[["PQ"]], true_dist, true_mean, true_sd) +
      draw_si(counts[["PQ"]], true_dist, true_mean, true_sd)
  }

  observed_icc <- pmax(round(c(cp, ps, pt, pq)), 0)

  data.frame(
    observed_icc = observed_icc,
    latent_path = rep(names(counts), counts),
    stringsAsFactors = FALSE
  )
}

fit_one_replicate <- function(scenario_row, replicate_id) {
  weights <- c(
    CP = scenario_row[["w_cp"]],
    PS = scenario_row[["w_ps"]],
    PT = scenario_row[["w_pt"]],
    PQ = scenario_row[["w_pq"]]
  )

  sim_dat <- simulate_icc_intervals(
    n_obs = scenario_row[["n_obs"]],
    true_dist = scenario_row[["true_dist"]],
    true_mean = scenario_row[["true_mean"]],
    true_sd = scenario_row[["true_sd"]],
    weights = weights
  )

  fit <- tryCatch(
    si_estim(
      dat = sim_dat$observed_icc,
      dist = "lognormal",
      n = n_iter,
      n_starts = n_starts
    ),
    error = function(e) e
  )

  if (inherits(fit, "error")) {
    return(data.frame(
      scenario_id = scenario_row[["scenario_id"]],
      replicate_id = replicate_id,
      scenario_type = scenario_row[["scenario_type"]],
      true_dist = scenario_row[["true_dist"]],
      n_obs = scenario_row[["n_obs"]],
      weights = scenario_row[["weights"]],
      true_mean = scenario_row[["true_mean"]],
      true_sd = scenario_row[["true_sd"]],
      est_mean = NA_real_,
      est_sd = NA_real_,
      bias_mean = NA_real_,
      bias_sd = NA_real_,
      converged = FALSE,
      iterations = NA_integer_,
      stringsAsFactors = FALSE
    ))
  }

  est_mean <- fit[["mean"]]
  est_sd <- fit[["sd"]]

  data.frame(
    scenario_id = scenario_row[["scenario_id"]],
    replicate_id = replicate_id,
    scenario_type = scenario_row[["scenario_type"]],
    true_dist = scenario_row[["true_dist"]],
    n_obs = scenario_row[["n_obs"]],
    weights = scenario_row[["weights"]],
    true_mean = scenario_row[["true_mean"]],
    true_sd = scenario_row[["true_sd"]],
    est_mean = est_mean,
    est_sd = est_sd,
    bias_mean = est_mean - scenario_row[["true_mean"]],
    bias_sd = est_sd - scenario_row[["true_sd"]],
    converged = isTRUE(fit[["converged"]]),
    iterations = fit[["iterations"]],
    stringsAsFactors = FALSE
  )
}

summarise_results <- function(results) {
  split_results <- split(results, results$scenario_id)

  summary_list <- lapply(split_results, function(dat) {
    data.frame(
      scenario_id = dat$scenario_id[1],
      scenario_type = dat$scenario_type[1],
      true_dist = dat$true_dist[1],
      n_obs = dat$n_obs[1],
      weights = dat$weights[1],
      true_mean = dat$true_mean[1],
      true_sd = dat$true_sd[1],
      mean_est_mean = mean(dat$est_mean, na.rm = TRUE),
      bias_mean = mean(dat$bias_mean, na.rm = TRUE),
      rmse_mean = sqrt(mean(dat$bias_mean^2, na.rm = TRUE)),
      emp_sd_mean = sd(dat$est_mean, na.rm = TRUE),
      mean_est_sd = mean(dat$est_sd, na.rm = TRUE),
      bias_sd = mean(dat$bias_sd, na.rm = TRUE),
      rmse_sd = sqrt(mean(dat$bias_sd^2, na.rm = TRUE)),
      emp_sd_sd = sd(dat$est_sd, na.rm = TRUE),
      convergence_rate = mean(dat$converged, na.rm = TRUE),
      stringsAsFactors = FALSE
    )
  })

  out <- do.call(rbind, summary_list)
  rownames(out) <- NULL
  out[match(unique(results$scenario_id), out$scenario_id), ]
}
```

# 5. Run Simulations

```{r run-simulations, cache=TRUE}
progress_line <- function(...) {
  cat(sprintf(...), "\n")
  flush.console()
}

overall_start <- Sys.time()
total_scenarios <- nrow(scenario_grid)
total_fits <- total_scenarios * n_reps
fit_counter <- 0L

scenario_results <- vector("list", total_scenarios)

for (i in seq_len(total_scenarios)) {
  scenario_row <- scenario_grid[i, ]
  scenario_start <- Sys.time()

  progress_line(
    "[%s] Scenario %d/%d: %s | true_dist=%s | n_obs=%s | reps=%d",
    format(Sys.time(), "%H:%M:%S"),
    i,
    total_scenarios,
    scenario_row[["scenario_id"]],
    scenario_row[["true_dist"]],
    scenario_row[["n_obs"]],
    n_reps
  )

  replicate_list <- vector("list", n_reps)

  for (rep_id in seq_len(n_reps)) {
    fit_counter <- fit_counter + 1L
    replicate_list[[rep_id]] <- fit_one_replicate(
      scenario_row,
      replicate_id = rep_id
    )

    should_report <- rep_id == 1L ||
      rep_id == n_reps ||
      (progress_every > 0 && rep_id %% progress_every == 0L)

    if (should_report) {
      scenario_elapsed <- round(
        as.numeric(difftime(Sys.time(), scenario_start, units = "mins")),
        2
      )
      overall_elapsed <- round(
        as.numeric(difftime(Sys.time(), overall_start, units = "mins")),
        2
      )
      avg_seconds_per_fit <- as.numeric(
        difftime(Sys.time(), overall_start, units = "secs")
      ) / fit_counter
      remaining_fits <- total_fits - fit_counter
      eta_minutes <- round(avg_seconds_per_fit * remaining_fits / 60, 2)

      progress_line(
        "[%s]   Replicate %d/%d for %s; total fits %d/%d; scenario elapsed %.2f min; overall elapsed %.2f min; avg %.2f sec/fit; ETA %.2f min",
        format(Sys.time(), "%H:%M:%S"),
        rep_id,
        n_reps,
        scenario_row[["scenario_id"]],
        fit_counter,
        total_fits,
        scenario_elapsed,
        overall_elapsed,
        avg_seconds_per_fit,
        eta_minutes
      )
    }
  }

  scenario_results[[i]] <- do.call(rbind, replicate_list)
}

replicate_results <- do.call(rbind, scenario_results)

head(replicate_results)
```
