---
title: "Estimation of time-varying reproduction number"
author: "Kylie Ainslie"
date: "`r Sys.Date()`"
bibliography: references.bib
csl: nature.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup, echo = FALSE, message=FALSE, results='hide'}
library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)
library(EpiLPS)
library(zoo)
library(outbreaks)
library(fdrtool)
library(lubridate)
library(gridExtra)
library(RColorBrewer)
library(devtools)
load_all()
```

# Introduction
The `mitey` package is a lightweight package designed to provide easy implementation of the methods used in [Ainslie et al. 2024](https://kylieainslie.github.io/mitey/articles/epidemiology_of_scabies.html) to estimate epidemiological characteristics of scabies transmission. However, these methods are more widely applicable than in the context of scabies. One of the key functionalities of `mitey` is the estimation of the time-varying case reproduction number using data on time of symptom onset. The case reproduction number ($R^c_t$) is defined as the average number of new infections that an individual who becomes infected, or symptomatic, at a particular time point will go on to cause [@gostic2020], and is useful in retrospective analyses. The method of Wallinga and Lipsitch estimates the time-varying case reproduction number by determining the likelihood of an event occurring for every pair of time points [@wallinga2007]. The method requires no assumptions beyond the specification of the serial interval distribution, making it straightforward and easy to implement. 

In this article, we will demonstrate how to use `mitey` to estimate $R^c_t$ using a synthetic data set and real data. We will also compare the estimates from `mitey` to estimates from other R packages that can be used for similar analyses, namely `EpiEstim` [@cori2013] and `EpiLPS` [@gressani2022].

## Mathematical Background

The Wallinga-Lipsitch method estimates the case reproduction number based on the idea that the relative likelihood that case j infected case i depends on their time difference, weighted by the serial interval distribution.

For a given case i that showed symptoms at time t_i, the probability that it was infected by case j (with symptom onset at time t_j) is:

$$p_{i,j} = \frac{w(t_i - t_j)}{\sum_{k} w(t_i - t_k)}$$

Where:
- $w(\tau)$ is the probability mass function of the serial interval for a delay of $\tau$
- The denominator sums over all potential infectors $k$

The reproduction number for cases with symptom onset at time $t_j$ is then:

$$R(t_j) = \frac{\sum_i I_i \cdot p_{i,j}}{I_j}$$

Where $I_i$ and $I_j$ are the number of cases at times $t_i$ and $t_j$, respectively.

## Function Documentation

### `wallinga_lipsitch()`

#### Description
Estimates the time-varying reproduction number using the Wallinga-Lipsitch method, which calculates the case reproduction number based on symptom onset data and serial interval distribution.

#### Usage
```R
wallinga_lipsitch(
  incidence,
  dates,
  si_mean,
  si_sd,
  si_dist = "gamma",
  smoothing = 0,
  bootstrap = FALSE,
  n_bootstrap = 1000,
  conf_level = 0.95,
  shift = FALSE
)
```

#### Arguments

- incidence: Numeric vector of daily case counts.
- dates: Vector of dates corresponding to the incidence data.
- si_mean: Mean of the serial interval distribution.
- si_sd: Standard deviation of the serial interval distribution.
- si_dist: Distribution to use for serial interval ("gamma" or "normal").
- smoothing: Window size for smoothing estimates (0 for no smoothing).
- bootstrap: Logical; whether to compute bootstrap confidence intervals.
- n_bootstrap: Number of bootstrap samples to generate.
- conf_level: Confidence level for intervals (0.95 = 95% CI).
- shift: Logical; whether to shift estimates by one mean serial interval.

#### Returns
A data frame with columns:

- date: Original dates from input
- incidence: Daily case counts
- R: Estimated case reproduction number
- R_corrected: Case reproduction number with right-truncation correction

If bootstrap=TRUE:

- R_lower, R_upper: Confidence intervals for R
- R_corrected_lower, R_corrected_upper: Confidence intervals for R_corrected

If shift=TRUE:

- shifted_date: Dates shifted forward by one mean serial interval

## Right-Truncation Correction

The Wallinga-Lipsitch method includes a correction for right-truncation, which addresses the bias in reproduction number estimates near the end of the time series. This bias occurs because some secondary cases infected by recent primary cases have not yet been observed due to the delay represented by the serial interval.

### Implementation

1. For each case with symptom onset at time t, we calculate how many days have passed between t and the end of the observation period.
2. Using the serial interval distribution, we calculate the probability that a secondary infection from this case would have been observed by the end of our observation period.
3. We then adjust the reproduction number estimate by dividing by this probability, giving more weight to recent cases that have had less time to generate observable secondary cases.

This correction becomes increasingly important for cases near the end of the time series, but can introduce additional variance. Therefore, estimates from the most recent days should still be interpreted with caution.

# Synthetic data
First, we will generate a synthetic time series of incidence data using the function `generate_synthetic_epidemic()` which uses a renewal equation to generate the case incidence data. The renewal equation used is: 
$$\lambda_t = \sum_{i=1}^{t-1} R_i \times I_i \times w(t-i)$$

Where:
- $\lambda_t$ = expected new cases on day $t$
- $R_i$ = reproduction number on day $i$
- $I_i$ = observed cases on day $i$
- $w(t-i)$ = probability mass of serial interval for delay $(t-i)$.

Before we generate the synthetic data we need to specify the true time-varying reproduction numbers and the parameters of the serial interval distribution (mean and standard deviation).

```{r generate_si_dist}

# Set seed for reproducibility
set.seed(123)

# True Rt values
true_r_values <- c(
  rep(1.5, 20),                            # Start with R=1.5
  rep(2.0, 20),                            # Increase to R=2.0
  rep(0.8, 20),                            # Decrease to R=0.8
  rep(1.2, 20)                             # Increase to R=1.2
)

# Set serial interval parameters
si_mean <- 5
si_sd <- 2
si_dist <- "gamma"

# Generate synthetic data with high initial cases for stability
synthetic_data <- generate_synthetic_epidemic(
  true_r = true_r_values,
  si_mean = si_mean,
  si_sd = si_sd,
  si_dist = si_dist,
  initial_cases = 500
)
```

```{r, echo=FALSE}
head(synthetic_data)
```

```{r plot_synth_epi, echo = FALSE, message=FALSE, warning=FALSE}
# 1. True R values over time
p1 <- ggplot(synthetic_data, aes(x = date, y = true_r)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(title = "True Reproduction Number",
       x = "Date", y = "Rt") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# 2. Incident cases over time
p2 <- ggplot(synthetic_data, aes(x = date, y = incidence)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  labs(title = "Daily Incidence",
       x = "Date", y = "Cases") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Combine plots in a 2x2 grid
grid.arrange(p1, p2, ncol = 1)
```

## Estimating the case reproduction number
Next, using the simulated incidence data, we can estimate the time-varying case reproduction number using the method developed by Wallinga and Lipsitch [@wallinga2007]. 

<!-- We will use the `rt_estim()` function within `mitey`. `rt_estim()` expects an input data set (`inc_dat`) with two columns `inc` (the number of new cases for each time point) and `onset_date` (the time or date in which the new cases occurred). `rt_estim()` also requires the specification of the mean and standard deviation of the serial interval distribution and the underlying serial interval distribution (it currently only accepts "normal" and "gamma"). -->

```{r estimate_rt, echo=TRUE}
results <- wallinga_lipsitch(
  incidence = synthetic_data$incidence,
  dates = synthetic_data$date,
  si_mean = si_mean,
  si_sd = si_sd,
  si_dist = "gamma",
  smoothing = 0,
  bootstrap = TRUE,
  n_bootstrap = 1000,
  conf_level = 0.95
)
```

```{r look_at_results, echo=FALSE}
head(results)
```

Now, we can compare the estimated $R^c_t$ values to the true $R_t$ that we specified. However, an important thing to note is that the "true" $R_t$ is the instantaneous reproduction number. We are estimating the case reproduction number, and therefore, our estimates should be shifted to the left by one serial interval. When we shift the estimates by a serial interval (blue dashed line), we see that our estimates are in agreement with the true $R_t$.

```{r plot_data_wrangling, echo=FALSE}
# Create shifted estimates (accounting for serial interval lag)
offset <- round(si_mean)
n_days <- length(results$date)

# Create vectors for shifted values
shifted_indices <- (1+offset):n_days
original_indices <- 1:(n_days-offset)

# Initialize shifted vectors with NAs
shifted_r <- rep(NA, n_days)
shifted_r_lower <- rep(NA, n_days)
shifted_r_upper <- rep(NA, n_days)

# Perform shifting
shifted_r[original_indices] <- results$R[shifted_indices]
shifted_r_lower[original_indices] <- results$R_lower[shifted_indices]
shifted_r_upper[original_indices] <- results$R_upper[shifted_indices]

# Add to results
results$R_shifted <- shifted_r
results$R_shifted_lower <- shifted_r_lower
results$R_shifted_upper <- shifted_r_upper

# Combine with true R
plot_data <- data.frame(
  date = synthetic_data$date,
  true_r = synthetic_data$true_r,
  incidence = synthetic_data$incidence,
  results
)
```

```{r plot_sim_rt, echo=FALSE}
ggplot(plot_data, aes(x = date)) +
  # geom_ribbon(aes(ymin = R_lower, ymax = R_upper), 
              # fill = "#21908C", alpha = 0.2) +
  # geom_ribbon(aes(ymin = R_shifted_lower, ymax = R_shifted_upper), 
  #             fill = "#21908C", alpha = 0.2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(aes(y = true_r, color = "True R"), linewidth = 1.2) +
  geom_line(aes(y = R, color = "Estimated R"), linewidth = 1) +
  geom_line(aes(y = R_shifted, color = "Shifted R"), linewidth = 1, linetype = "dotted") +
  scale_y_continuous(limits = c(0, 5)) +
  # Customize colors
  scale_color_manual(name = "R Values", 
                    values = c("True R" = "black", 
                              "Estimated R" = "#21908C", 
                              "Shifted R" = "#21908C")) +
  # Add labels
  labs(x = "Date", y = "Reproduction Number (Rt)") +
  # Customize theme
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Sensitivity to Serial Interval Parameters

The estimates produced by the Wallinga-Lipsitch method are sensitive to the specified serial interval distribution. Let's explore how changes in the mean and standard deviation affect our estimates for the synthetic epidemic:

```{r sensitivity_analysis, echo=FALSE, fig.width=9, fig.height=6}
# Basic parameter set
base_mean <- 5
base_sd <- 2

# Alternative parameter sets
param_sets <- list(
  list(mean = 4, sd = 2, label = "Mean = 4, SD = 2"),
  list(mean = 6, sd = 2, label = "Mean = 6, SD = 2"),
  list(mean = 5, sd = 1, label = "Mean = 5, SD = 1"),
  list(mean = 5, sd = 3, label = "Mean = 5, SD = 3")
)

# Calculate R for each parameter set
sensitivity_results <- list()

for (i in 1:length(param_sets)) {
  params <- param_sets[[i]]
  
  result <- wallinga_lipsitch(
    incidence = synthetic_data$incidence,
    dates = synthetic_data$date,
    si_mean = params$mean,
    si_sd = params$sd,
    si_dist = "gamma",
    bootstrap = FALSE
  )
  
  # Store results with parameter label
  result$parameter_set <- params$label
  sensitivity_results[[i]] <- result
}

# Combine results
all_sensitivity <- do.call(rbind, sensitivity_results)

true_r_df <- data.frame(
  date = synthetic_data$date, 
  R = synthetic_data$true_r,
  parameter_set = "True R"
)

plot_data <- rbind(
  all_sensitivity[, c("date", "R", "parameter_set")],
  true_r_df
)

# Plot with custom colors including black for True R
ggplot(plot_data, aes(x = date, y = R, color = parameter_set, size = parameter_set)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  scale_color_manual(name = "Serial Interval Parameters",
                    values = c(brewer.pal(4, "Set1"), "black")) +
  scale_size_manual(values = c(rep(1, 4), 1.2), guide = "none") +
  labs(x = "Date", y = "Reproduction Number (R)") +
  theme_minimal() +
  theme(
    legend.position.inside = c(0.95, 0.95),    # x and y coordinates (0-1 range)
    legend.justification = c(1, 1),     # anchor point (top-right corner)
    legend.background = element_rect(fill = "white", color = NA, linewidth = 0.5),
    legend.margin = margin(6, 6, 6, 6)  # add some margin inside the legend box
  )
```

## Comparing methods
Using the `estimRmcmc` function within the `EpiLPS` package, we can compare our estimates to those produced my `EpiLPS` and `EpiEstim` (by specifying `Cori = TRUE`) for estimating the instantaneous reproduction number. We can also estimate the case reproduction number using the method of Wallinga and Teunis [@wallinga2004] as estimated by `EpiEstim` (by specifying `WTR = TRUE`). 

```{r compare_methods, results='hide'}
si_spec <- Idist(mean = si_mean, sd = si_sd, dist = si_dist)

fitmcmc <- estimRmcmc(incidence = synthetic_data$incidence, si = si_spec$pvec,
                      CoriR = TRUE, WTR = TRUE,
                      niter = 5000, burnin = 2000)
```

```{r summary_fit, warning=FALSE, message=FALSE}
summary(fitmcmc)
```

```{r comparison_plot_wrangling, echo=FALSE}
comparison_df <- data.frame(
  time = 1:length(synthetic_data$true_r),
  true_r = synthetic_data$true_r,
  
  # Your Wallinga-Lipsitch method
  wl_r = results$R,
  wl_r_lower = results$R_lower,
  wl_r_upper = results$R_upper,
  
  # Shifted WL
  wl_shifted = shifted_r,
  wl_shifted_lower = shifted_r_lower,
  wl_shifted_upper = shifted_r_upper
)

# Add the results from fitmcmc
# Make sure lengths match by trimming or padding as needed
comparison_df$epilps_r <- c(fitmcmc$RLPS$R, rep(NA, max(0, nrow(comparison_df) - length(fitmcmc$RLPS$R))))
comparison_df$epilps_lower <- c(fitmcmc$RLPS$Rq0.025, rep(NA, max(0, nrow(comparison_df) - length(fitmcmc$RLPS$Rq0.025))))
comparison_df$epilps_upper <- c(fitmcmc$RLPS$Rq0.975, rep(NA, max(0, nrow(comparison_df) - length(fitmcmc$RLPS$Rq0.975))))

# Add EpiEstim (Cori) results
if (length(fitmcmc$RCori$`Mean(R)`) < nrow(comparison_df)) {
  # Pad with NAs if shorter
  cori_r <- c(fitmcmc$RCori$`Mean(R)`, rep(NA, nrow(comparison_df) - length(fitmcmc$RCori$`Mean(R)`)))
  cori_lower <- c(fitmcmc$RCori$`Quantile.0.025(R)`, rep(NA, nrow(comparison_df) - length(fitmcmc$RCori$`Quantile.0.025(R)`)))
  cori_upper <- c(fitmcmc$RCori$`Quantile.0.975(R)`, rep(NA, nrow(comparison_df) - length(fitmcmc$RCori$`Quantile.0.975(R)`)))
} else {
  # Trim if longer
  cori_r <- fitmcmc$RCori$`Mean(R)`[1:nrow(comparison_df)]
  cori_lower <- fitmcmc$RCori$`Quantile.0.025(R)`[1:nrow(comparison_df)]
  cori_upper <- fitmcmc$RCori$`Quantile.0.975(R)`[1:nrow(comparison_df)]
}

comparison_df$cori_r <- cori_r
comparison_df$cori_lower <- cori_lower
comparison_df$cori_upper <- cori_upper

# Add Wallinga-Teunis results
if (length(fitmcmc$RWT$`Mean(R)`) < nrow(comparison_df)) {
  # Pad with NAs if shorter
  wt_r <- c(fitmcmc$RWT$`Mean(R)`, rep(NA, nrow(comparison_df) - length(fitmcmc$RWT$`Mean(R)`)))
  wt_lower <- c(fitmcmc$RWT$`Quantile.0.025(R)`, rep(NA, nrow(comparison_df) - length(fitmcmc$RWT$`Quantile.0.025(R)`)))
  wt_upper <- c(fitmcmc$RWT$`Quantile.0.975(R)`, rep(NA, nrow(comparison_df) - length(fitmcmc$RWT$`Quantile.0.975(R)`)))
} else {
  # Trim if longer
  wt_r <- fitmcmc$RWT$`Mean(R)`[1:nrow(comparison_df)]
  wt_lower <- fitmcmc$RWT$`Quantile.0.025(R)`[1:nrow(comparison_df)]
  wt_upper <- fitmcmc$RWT$`Quantile.0.975(R)`[1:nrow(comparison_df)]
}

comparison_df$wt_r <- wt_r
comparison_df$wt_lower <- wt_lower
comparison_df$wt_upper <- wt_upper

# Cap values at 5 for better visualization
comparison_df <- comparison_df %>%
  mutate(across(contains(c("_r", "true", "lower", "upper")), ~pmin(., 5, na.rm = TRUE)))
```

```{r plot_comparison_methods, echo=FALSE, out.width = "100%"}
my_colors <- c("black", "#FDE725", "#21908C", "#440154", "#5DC863")

p <- ggplot(comparison_df %>%
              filter(time > min(time) + 7), aes(x = time)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_ribbon(aes(ymin = wl_r_lower, ymax = wl_r_upper), fill = my_colors[3], alpha = 0.1) +
  geom_ribbon(aes(ymin = cori_lower, ymax = cori_upper), fill = my_colors[5], alpha = 0.1) +
  geom_ribbon(aes(ymin = wt_lower, ymax = wt_upper), fill = my_colors[4], alpha = 0.1) +
  geom_ribbon(aes(ymin = epilps_lower, ymax = epilps_upper), fill = my_colors[2], alpha = 0.1) +
  
  geom_line(aes(y = true_r, color = "True R"), linewidth = 1.2) +
  geom_line(aes(y = epilps_r, color = "EpiLPS"), linewidth = 0.9) +
  geom_line(aes(y = cori_r, color = "EpiEstim"), linewidth = 0.9) +
  geom_line(aes(y = wt_r, color = "Wallinga-Teunis"), linewidth = 0.9) +
  geom_line(aes(y = wl_r, color = "Wallinga-Lipsitch"), linewidth = 0.9) +
  #geom_line(aes(y = wl_shifted, color = "Wallinga-Lipsitch"), linewidth = 0.9) +

  scale_y_continuous(limits = c(0, 5)) +
  scale_color_manual(name = "Method", 
                    values = c("True R" = my_colors[1], 
                              "EpiLPS" = my_colors[2],
                              "EpiEstim" = my_colors[5],
                              "Wallinga-Teunis" = my_colors[4],
                              "Wallinga-Lipsitch" = my_colors[3])) +

  labs(x = "Time", y = "Reproduction Number") +
  theme_minimal() +
  theme(legend.position = "bottom")

p
```

# Real-world case studies

## 1. Zika outbreak in Giradot, Colombia (2015)
To illustrate how to apply `wallinga_lipsitch()` to real data and compare it to other methods, we'll use data on daily incidence of the Zika virus disease in Giradot, Colombia from October 2015 to January 2016. The data is available from the `outbreaks` package and is called `zika2015`. 

```{r zika_data}
lapply(zika2015, head, 10)
```

```{r zika_epicurve}
zika_epicurve <- epicurve(zika2015$incidence, dates = zika2015$date, datelab = "14d")
zika_epicurve
```

Now we'll estimate the case reproduction number using `wallinga_lipsitch()`.

```{r use_mitey}
res_zika <- wallinga_lipsitch(
  incidence = zika2015$incidence,
  dates = zika2015$date,
  si_mean = 7,
  si_sd = 1.5,
  si_dist = "gamma",
  smoothing = 0,
  bootstrap = TRUE,
  n_bootstrap = 1000,
  conf_level = 0.95,
  shift = TRUE
)
```

```{r zika_plot, echo=FALSE}
# Create the plot
r_plot <- ggplot(res_zika %>%
         filter(date > min(date) + 7), 
       aes(x = date)) +
  geom_ribbon(aes(ymin = R_lower, ymax = R_upper), fill = "#21908C", alpha = 0.2) +
  geom_line(aes(y = R), color="#21908C", linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  coord_cartesian(ylim = c(0,8)) +
  scale_x_date(date_breaks = "7 days", date_labels = "%d-%m-%Y") +
  scale_color_manual(values = c("WL" = "#21908C")) +  
  scale_fill_manual(values = c("WL" = "#21908C")) + 
  labs(x = "Time", y = "R", title = "Estimated R") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

# show Rt plot with epicurve
gridExtra::grid.arrange(zika_epicurve, r_plot, nrow = 2, ncol = 1)
```

### Compare with EpiLPS and EpiEstim

Below we compare the estimates using the Walling and Lipsitch method (WL) with those of `EpiLPS` and the Wallinga and Teunis method (WT) produced by `EpiEstim`. We see similar results and see the expected shif in the estimates of EpiLPS because it is estimating the instantaneous reproduction number. Additionally, it is important to note that the below WL estimates have not been smoothed.

```{r use_epilps, warning=FALSE}
si <- Idist(mean = 7, sd = 1.5)
epifit <- estimR(zika2015$incidence, dates = zika2015$dates, si = si$pvec, WTR = TRUE)
```

```{r zika_comparison_plot, echo=FALSE}
r_plot_comparison <- ggplot(res_zika %>%
         filter(date > min(date) + 7), 
       aes(x = date)) +
  geom_ribbon(aes(ymin = R_lower, ymax = R_upper), fill = "#21908C", alpha = 0.2) +
  geom_line(aes(color = "WL"), y = res_zika$R[res_zika$date > min(res_zika$date) + 7], linewidth = 1) +  # Add color in aes()
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(data = epifit$RLPS[-c(1:7),], aes(x = Time, y = R, color = "EpiLPS"), linewidth = 1) +
  geom_line(data = epifit$RWT, aes(x = t_end + zika2015$dates[1], y = `Mean(R)`, color = "WT (EpiEstim)"), linewidth = 1) +
  coord_cartesian(ylim = c(0,8)) +
  scale_x_date(date_breaks = "7 days", date_labels = "%d-%m-%Y") +
  scale_color_manual(values = c("WL" = "#21908C", "EpiLPS" = "#FDE725", "WT (EpiEstim)" = "#440154")) + 
  labs(x = "Time", y = "R", title = "Estimated R", color = "Method") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = c(0.95, 0.95),  # Position in the top right
        legend.justification = c(1, 1),   # Align the top-right corner of the legend
        legend.background = element_rect(fill = "white", color = NA),
        legend.margin = margin(5, 5, 5, 5))
r_plot_comparison
```

## 2. Influenza A (H7N9) Outbreak in China (2013)
Again, we'll use a dataset from the `outbreaks` package. For this analysis, we'll use data from the 2013 influenza A (H7N9) outbreak in China.

```{r load_h7n9_data, message=FALSE, warning=FALSE}
data(fluH7N9_china_2013)
str(fluH7N9_china_2013)
```

The dataset contains details about 136 confirmed cases of H7N9 influenza in China between February and May 2013.

```{r prepare_h7n9_data, echo=FALSE}
# Extract relevant columns and create a daily incidence time series
h7n9_data <- fluH7N9_china_2013 %>%
  mutate(date_onset = as.Date(date_of_onset)) %>%
  filter(!is.na(date_onset)) %>%  # Remove cases with missing onset date
  group_by(date_onset) %>%
  dplyr::summarize(cases = n()) %>%
  arrange(date_onset)

# Create a complete time series with zeros for days with no cases
date_range <- seq(min(h7n9_data$date_onset), max(h7n9_data$date_onset), by = "day")
h7n9_complete <- data.frame(date_onset = date_range) %>%
  left_join(h7n9_data, by = "date_onset") %>%
  mutate(cases = ifelse(is.na(cases), 0, cases))

# Plot the epidemic curve
ggplot(h7n9_complete, aes(x = date_onset, y = cases)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Date",
    y = "Cases"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

The H7N9 influenza outbreak shows a clear epidemic pattern, with most cases occurring between March and April 2013. Based on the literature, the serial interval for influenza A (H7N9) has been estimated to be around 7-9 days.

Let's now estimate the time-varying reproduction number using the Wallinga-Lipsitch method:

```{r h7n9_rt_estimation, message=FALSE, warning=FALSE}
# Estimate Rt using the Wallinga-Lipsitch method
# For H7N9, we'll use serial interval estimates from the literature
# Mean: 8.1 days, SD: 3.4 days (based on Cowling et al., 2013)

h7n9_rt <- wallinga_lipsitch(
  incidence = h7n9_complete$cases,
  dates = h7n9_complete$date_onset,
  si_mean = 8.1,
  si_sd = 3.4,
  si_dist = "gamma",
  smoothing = 7,  # Apply 7-day smoothing for weekly patterns
  bootstrap = TRUE,
  n_bootstrap = 500,
  conf_level = 0.95,
  shift = TRUE
)
```

```{r h7n9_plot, echo=FALSE}
# Create a combined plot with incidence and Rt
p1 <- ggplot(h7n9_complete, aes(x = date_onset, y = cases)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Daily incidence",
    x = NULL,
    y = "Number of cases"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- ggplot(h7n9_rt %>% filter(!is.na(R)), aes(x = date)) +
  geom_ribbon(aes(ymin = R_lower, ymax = R_upper), fill = "orange", alpha = 0.3) +
  geom_line(aes(y = R), color = "darkorange", linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red") +
  labs(
    title = "Reproduction number",
    x = "Date",
    y = "Rt"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

# Combine the plots using gridExtra
gridExtra::grid.arrange(p1, p2, nrow = 2, heights = c(1, 1.5))

```

To provide a comprehensive methodological comparison, let's compare our Wallinga-Lipsitch implementation with the `EpiLPS` and `EpiEstim` packages, similar to our Zika analysis:

```{r h7n9_compare_methods, message=FALSE, warning=FALSE, echo=FALSE}
# Create serial interval distribution for EpiLPS and EpiEstim
h7n9_si <- Idist(mean = 8.1, sd = 3.4, dist = "gamma")

# Get estimates from EpiLPS and EpiEstim (Wallinga-Teunis)
h7n9_epifit <- estimR(
  h7n9_complete$cases, 
  dates = h7n9_complete$date_onset, 
  si = h7n9_si$pvec, 
  WTR = TRUE
)

# Create a base Rt plot using our Wallinga-Lipsitch method
h7n9_base_plot <- ggplot(h7n9_rt %>% 
                         filter(!is.na(R) & date >= as.Date("2013-02-25")), 
                       aes(x = date)) +
  geom_ribbon(aes(ymin = R_lower, ymax = R_upper, fill = "WL"), alpha = 0.3) +
  geom_line(aes(y = R, color = "WL"), linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
  # geom_vline(xintercept = intervention_date, linetype = "dotted", color = "darkred") +
  # annotate("text", x = intervention_date + 3, y = 4, 
  #          label = "Market closures", hjust = 0, size = 3, color = "darkred") +
  coord_cartesian(ylim = c(0,4)) +
  labs(
    x = "Date", 
    y = "Reproduction Number",
    title = "Comparison of Rt Estimation Methods",
    subtitle = "H7N9 Influenza Outbreak, China 2013"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 9)
  )

# Add results from other methods
h7n9_comparison <- h7n9_base_plot +
  # Add EpiLPS line
  geom_line(data = h7n9_epifit$RLPS[-c(1:7),], 
           aes(x = Time, y = R, color = "EpiLPS"), 
           linewidth = 1, linetype = "solid") +
  # Add Wallinga-Teunis line from EpiEstim
  geom_line(data = h7n9_epifit$RWT, 
           aes(x = t_end + h7n9_complete$date_onset[1], y = `Mean(R)`, color = "WT (EpiEstim)"), 
           linewidth = 1, linetype = "solid") +
  # Set colors for the methods
  scale_color_manual(
    name = "Method",
    values = c("WL" = "#21908C", "EpiLPS" = "#FDE725", "WT (EpiEstim)" = "#440154")
  ) +
  scale_fill_manual(
    name = "Method",
    values = c("WL" = "#21908C")
  ) +
  # Place legend at the bottom
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  ) +
  guides(fill = "none")  # Hide the fill from the legend

h7n9_comparison
```

The comparison reveals several interesting methodological insights when applied to the H7N9 outbreak:

1. **Timing differences**: The Wallinga-Lipsitch (WL) method and Wallinga-Teunis (WT) method from EpiEstim produce roughly similar temporal patterns, but with some timing differences. This is expected as they both estimate the case reproduction number but use slightly different mathematical approaches.

2. **EpiLPS vs. WL/WT**: The EpiLPS method, which estimates the instantaneous reproduction number, shows a different pattern, particularly in terms of when changes in transmission are detected. This is consistent with the theoretical differences between instantaneous and case reproduction numbers.

3. **Uncertainty handling**: Note that our WL method includes confidence intervals (shown as the blue shaded region), providing important context about estimation uncertainty that should be considered when interpreting results.

# References

<!-- # Checks  -->
<!-- To sense-check estimates of the reproduction number from real data, there are a couple of strategies : -->
<!--   1. Growth rate method: calculate $R_t = b(t+u)/b(t)$ where $b()$ is the incidence, t is the current time point, and u is the mean generation time (here, approximated with the mean serial interval). -->

<!-- ```{r rt_growth_rate_method, eval=FALSE} -->
<!--   # 1. Calculate b(t+u)/b(t) where b() is the incidence, t is the current time -->
<!-- #    point, and u is the mean SI -->
<!-- seq1 <- 1:(nrow(df_real)) -->
<!-- seq2 <- seq1 + 123 -->
<!-- rt_check1 <- data.frame( -->
<!--   onset_date = (seq1 - 1) + min(df$onset_date), -->
<!--   rt_check1 = df_real$inc[seq2]/df_real$inc[seq1] -->
<!-- )  -->
<!-- ``` -->

<!--   2. Calculate the geometric mean of the time series of time-varying reproduction number. The geometric mean should be ~$R_0$. -->
<!-- ```{r rt_geom_mean, eval=FALSE}   -->
<!-- # 2. Calculate geometric mean -->
<!-- # This should be approx R0 = 1.1 -->
<!-- rt_check2 <- exp(mean(log(rt_estimates1$rt), na.rm = TRUE)) -->
<!-- rt_check2 -->
<!-- ```   -->

