---
title: "Validation of time-varying reproduction number estimation"
author: "Kylie Ainslie"
date: "`r Sys.Date()`"
bibliography: references.bib
csl: nature.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, message=FALSE, results='hide'}
library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)
library(EpiLPS)
library(zoo)
library(outbreaks)
library(devtools)
load_all()
```

# Introduction
The `mitey` package is a lightweight package designed to provide easy implementation of the methods used in [Ainsie et al. 2024](https://kylieainslie.github.io/mitey/articles/epidemiology_of_scabies.html) to estimate epidemiological characteristics of scabies transmission. However, these methods are more widely applicable than in the context of scabies. One of the key functionalities of `mitey` is the estimation of the time-varying case reproduction number using data on time of symptom onset. The case reproduction number ($R^c_t$) is defined as the average number of new infections that an individual who becomes infected, or symptomatic, at a particular time point will go on to cause [@gostic2020], and is useful in retrospective analyses. The method of Wallinga and Lipsitch estimates the time-varying case reproduction number by determining the likelihood of an event occurring for every pair of time points [@wallinga2007]. The method requires no assumptions beyond the specification of the serial interval distribution, making it straightforward and easy to implement. 

In this article, we will demonstrate how to use `mitey` to estimate $R^c_t$ using a synthetic data set and real data. We will also compare the estimates from `mitey` to estimates from other R packages that can be used for similar analyses, namely `EpiEstim` [@cori2013] and `EpiLPS` [@gressani2022].

# Synthetic data
First, we will generate a synthetic time series of incidence data using the `episim()` function from `EpiLPS` which uses a renewal equation model to generate the case incidence data based on a Poisson or negative binomial process [@gressani2022]. For full details see [here](https://doi.org/10.1371/journal.pcbi.1010618.s002). `episim()` requires the specification of the (discrete) serial interval distribution, which can be specified using the `Idist()` function.

```{r generate_si_dist}
# specify serial interval distribution
mean_si <- 8
sd_si <- 5.66

# generate discrete serial interval distribution using EpiLPS::Idist
si_spec <- Idist(mean = mean_si, sd = sd_si, dist = "gamma")
plot(si_spec, titlesize = 12)
```

Now, using `episim()` we can generate incidence data by specifying the generated serial interval distribution, the functional form of the reproduction number (here `Rpattern = 5`), and the length of the epidemic (here 40 days).

```{r simulate_inc_data}
# set seed
set.seed(1234)

# generate data
t_end <- 40
datasim <- episim(si = si_spec$pvec, Rpattern = 5, endepi = t_end,
                  dist = "negbin", overdisp = 15)

epicurve(datasim$y, title = "Simulated epidemic curve")

```

# Estimating the case reproduction number
Next, using the simulated incidence data `datasim$y`, we can estimate the time-varying case reproduction number using the method developed by Wallinga and Lipsitch [@wallinga2007]. We will use the `rt_estim()` function within `mitey`. `rt_estim()` expects an input data set (`inc_dat`) with two columns `inc` (the number of new cases for each time point) and `onset_date` (the time or date in which the new cases occurred). `rt_estim()` also requires the specification of the mean and standard deviation of the serial interval distribution and the underlying serial interval distribution (it currently only accepts "normal" and "gamma").

```{r estimate_rt, echo=TRUE}
# Estimate Rt with mitey::rt_estim()

t_vec <- seq(1, length(datasim$y), by = 1)
inc_dat <- data.frame(onset_date = t_vec, inc = datasim$y)

rt_estimated <- rt_estim(
  inc_dat = inc_dat,
  mean_si = mean_si,
  sd_si = sd_si,
  dist_si = "gamma"
)
```

```{r look_at_results, echo=FALSE}
head(rt_estimated)
```

Now, we can compare the estimated $R^c_t$ values (solid blue line) to the true $R_t$ that is outputted by `episim()`, here `datasim$Rtrue`. However, an important thing to note is that the "true" $R_t$ is the instantaneous reproduction number. We are estimating the case reproduction number, and therefore, our estimates should be shifted to the left by one serial interval. When we shift the estimates by a serial interval (blue dashed line), we see that our estimates are in agreement with the true $R_t$.

```{r plot_results, echo=FALSE}
# time and true Rt vectors
tt <- seq(mean_si + 1, t_end, by = 1)
Rtrue <- sapply(tt, datasim$Rtrue)

# clean-up the estimates a little
rt_WL <- ifelse(is.nan(rt_estimated$rt), NA, rt_estimated$rt)
rt_WL <- ifelse(is.infinite(rt_estimated$rt) | rt_estimated$rt > 15, NA, rt_estimated$rt)

# smooth and shift estimated Rt
rt_WL_smooth <- rollmean(rt_WL, k = 7, fill = NA, na.rm = TRUE)
rt_WL_shift <- c(rep(NA, mean_si), rt_WL_smooth)

# plot
plot(tt, Rtrue, type = "l", xlab = "Time", ylab = "R", ylim = c(0,4),
     lwd = 2)
lines(tt, rt_WL_smooth[tt], col = "blue", lwd = 2)
lines(tt, rt_WL_shift[tt], col = "blue", lty = 2, lwd = 2)
legend("topright", col = c("black","blue","blue"),
       c("True R","WL", "WL (shifted)"),
       bty = "n", lty = c(1,1,2))

```

To obtain confidence bounds for our estimated $R_t$ values, we can use `rt_estim_w_boot()` and specify the number of bootstrap samples using `n_bootstrap = ` argument. 

```{r rt_boot, eval = FALSE}
rt_estimated_boot <- rt_estim_w_boot(
  inc_dat = inc_dat,
  mean_si = gi_mean,
  sd_si = gi_sd,
  dist_si = "gamma",
  n_bootstrap = 100
)
```

# Comparing methods
Using the `estimRmcmc` function within the `EpiLPS` package, we can compare our estimates to those produced my `EpiLPS` and `EpiEstim` (by specifying `Cori = TRUE`) for estimating the instantaneous reproduction number. We can also estimate the case reproduction number using the method of Wallinga and Teunis [@wallinga2004] as estimated by `EpiEstim` (by specifying `WTR = TRUE`).

```{r compare_methods, results='hide'}
fitmcmc <- estimRmcmc(incidence = datasim$y, si = si_spec$pvec,
                      CoriR = TRUE, WTR = TRUE,
                      niter = 5000, burnin = 2000)
```

```{r summary_fit}
summary(fitmcmc)
```

```{r plot_comparison_methods, out.width = "100%"}
plot(tt, Rtrue, type = "l", xlab = "Time", ylab = "R", ylim = c(0,4),
     lwd = 2)
lines(tt, fitmcmc$RLPS$R[-(1:mean_si)], col = "red", lwd = 2)
lines(tt, fitmcmc$RCori$`Mean(R)`, col = "green", lwd = 2)
lines(tt, fitmcmc$RWT$`Mean(R)`[-1], col = "purple", lwd = 2)
lines(tt, rt_WL_smooth[tt], col = "blue", lwd = 2)
lines(tt, rt_WL_shift[tt], col = "blue", lty = 2, lwd = 2)
legend("topright", col = c("black","red","green", "purple", "blue", "blue"),
       c("True R","EpiLPS","EpiEstim", "WT", "WL", "WL (shifted)"),
       bty = "n", lty = c(1,1,1,1,1,2))
```

# Real data

## Zika outbreak in Giradot, Colombia (2015)
To illustrate how to apply `rt_estim()` to real data and compare it to other methods, we'll use data on daily incidence of the Zika virus disease in Giradot, Colombia from October 2015 to January 2016. The data is available from the `outbreaks` package and is called `zika2015`. 

```{r zika_data}
lapply(zika2015, head, 10)
```

```{r zika_epicurve}
zika_epicurve <- epicurve(zika2015$incidence, dates = zika2015$date, datelab = "14d")
zika_epicurve
```

Now we'll estimate the case reproduction number using `rt_estim()`.

```{r use_mitey}
# Estimate Rt with mitey::rt_estim()
inc_dat <- data.frame(onset_date = zika2015$dates, inc = zika2015$incidence)

rt_estimated <- rt_estim_w_boot(
  inc_dat = inc_dat,
  mean_si = 7,
  sd_si = 1.5,
  dist_si = "gamma",
  n_bootstrap = 100
)
```

```{r plot}
# Create the plot
r_plot <- ggplot(rt_estimated$results %>%
         filter(onset_date > min(onset_date) + 7,
                !is.na(median_rt_adjusted)), 
       aes(x = onset_date)) +
  geom_ribbon(aes(ymin = lower_rt_adjusted, ymax = upper_rt_adjusted), fill = "#21908C", alpha = 0.2) +
  geom_line(aes(y = median_rt_adjusted, color = "WL"), size = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black") +
  coord_cartesian(ylim = c(0,12)) +
  scale_x_date(date_breaks = "7 days", date_labels = "%d-%m-%Y") +
  scale_color_manual(values = c("WL" = "#21908C")) +  
  scale_fill_manual(values = c("WL" = "#21908C")) + 
  labs(x = "Time", y = "R", title = "Estimated R") +
  guides(color = "none", fill = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

# show Rt plot with epicurve
gridExtra::grid.arrange(zika_epicurve, r_plot, nrow = 2, ncol = 1)
```
### Compare with EpiLPS and EpiEstim
Below we compare the estimates using the Walling and Lipsitch method (WL) with those of `EpiLPS` and the Wallinga and Teunis method (WT) produced by `EpiEstim`. We see similar results and see the expected shif in the estimates of EpiLPS because it is estimating the instantaneous reproduction number. Additionally, it is important to note that the below WL estimates have not been smoothed.

```{r use_epilps, warning=FALSE}
si <- Idist(mean = 7, sd = 1.5)
epifit <- estimR(zika2015$incidence, dates = zika2015$dates, si = si$pvec, WTR = TRUE)

r_plot +
  geom_line(data = epifit$RLPS[-c(1:7),], aes(x = Time, y = R, color = "EpiLPS"), size = 1, linetype = "solid") +
  geom_line(data = epifit$RWT, aes(x = t_end + zika2015$dates[1], y = `Mean(R)`, color = "WT (EpiEstim)"), size = 1, linetype = "solid") +
  scale_color_manual(values = c("WL" = "#21908C", "EpiLPS" = "#FDE725", "WT (EpiEstim)" = "#440154")) + 
  labs(color = "Method")

```

# References

<!-- # Checks  -->
<!-- To sense-check estimates of the reproduction number from real data, there are a couple of strategies : -->
<!--   1. Growth rate method: calculate $R_t = b(t+u)/b(t)$ where $b()$ is the incidence, t is the current time point, and u is the mean generation time (here, approximated with the mean serial interval). -->

<!-- ```{r rt_growth_rate_method, eval=FALSE} -->
<!--   # 1. Calculate b(t+u)/b(t) where b() is the incidence, t is the current time -->
<!-- #    point, and u is the mean SI -->
<!-- seq1 <- 1:(nrow(df_real)) -->
<!-- seq2 <- seq1 + 123 -->
<!-- rt_check1 <- data.frame( -->
<!--   onset_date = (seq1 - 1) + min(df$onset_date), -->
<!--   rt_check1 = df_real$inc[seq2]/df_real$inc[seq1] -->
<!-- )  -->
<!-- ``` -->

<!--   2. Calculate the geometric mean of the time series of time-varying reproduction number. The geometric mean should be ~$R_0$. -->
<!-- ```{r rt_geom_mean, eval=FALSE}   -->
<!-- # 2. Calculate geometric mean -->
<!-- # This should be approx R0 = 1.1 -->
<!-- rt_check2 <- exp(mean(log(rt_estimates1$rt), na.rm = TRUE)) -->
<!-- rt_check2 -->
<!-- ```   -->

